# Data Sources {#sec-data-sources}

```{r}
#| label: setup
#| include: false
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
```

## Primary Data Sources

```{r}
#| label: tbl-datasources
#| tbl-cap: "Primary data sources used in the IMPT pipeline"

sources <- data.frame(
    Code = c("CAOP", "BGRI21", "IMOB", "GTFS", "OSM-RN", "OSM-POI", "CM", "ANSR", "INE-Inc", "INE-Hous", "GBA", "PMUS", "COS", "IP"),
    Name = c(
        "CAOP 2024.1 — Administrative Boundaries",
        "Census 2021 — Base Geográfica de Referenciação de Informação",
        "IMOB 2017/18 — Inquérito à Mobilidade nas Áreas Metropolitanas",
        "GTFS — General Transit Feed Specification (all AML operators)",
        "OpenStreetMap (OSM) — Road network",
        "OpenStreetMap (OSM) — POIs",
        "Carris Metropolitana — Health centres and schools datasets",
        "ANSR — Road accident data (Sinistralidade Rodoviária)",
        "INE — Household income by parish (Rendimento agregado 2023)",
        "INE — Housing costs by parish (Habitação custos 2021)",
        "Global Building Atlas - Building heights",
        "TML PMUS - Shared mobility docks",
        "COS 2023 v1 — Land use Map",
        "Infraestruturas de Portugal — Toll cost"
    ),
    Year = c("2024", "2021", "2018", "2026-02", "2026-01", "2024-04", "2024", "2019–2023", "2023", "2021", "2026", "2025", "2023", "2026"),
    URL = c(
        "https://www.dgterritorio.gov.pt/sites/default/files/ficheiros-cartografia/CAOP_Continente_2024.1-gpkg.zip",
        "https://mapas.ine.pt/download/filesGPG/2021/nuts3/BGRI21_170.zip",
        "https://www.ine.pt/xurl/pub/349495406",
        "Multiple operators — see <a href='#sec-gtfs-operators'>GTFS Operators</a>",
        "https://export.hotosm.org/exports/4782f0b8-6778-4c0e-8e4f-97fc62e7f240",
        "https://export.hotosm.org / https://github.com/U-Shift/SiteSelection/releases/download/0.1/osm_poi_landuse.gpkg",
        "https://github.com/carrismetropolitana/datasets",
        "https://pmmus.tmlmobilidade.pt/",
        "https://www.ine.pt/xurl/pub/66303599",
        "https://www.ine.pt/xurl/pub/65586079",
        "https://github.com/zhu-xlab/GlobalBuildingAtlas",
        "https://pmmus.tmlmobilidade.pt/",
        "https://geo2.dgterritorio.gov.pt/cos/S2/COS2023/COS2023v1-S2-gpkg.zip",
        "https://portagens.infraestruturasdeportugal.pt"
    ),
    stringsAsFactors = FALSE
)

# Convert URLs to clickable HTML links
sources$URL <- ifelse(grepl("^http", sources$URL),
    paste0("<a href=\"", sources$URL, "\" target=\"_blank\">", sources$URL, "</a>"),
    sources$URL
)

kbl(sources,
    col.names = c("Code", "Dataset", "Reference period", "Source"),
    booktabs = TRUE, longtable = TRUE, escape = FALSE
) |>
    kable_styling(font_size = 9, full_width = FALSE) |>
    column_spec(2, width = "9cm") |>
    column_spec(4, width = "5.5cm")
```

## GTFS Transit Operators {#sec-gtfs-operators}

Travel time matrices for public transit and stop-level frequency statistics were computed using GTFS feeds downloaded on **04/02/2026** from the following operators:

```{r}
#| label: tbl-gtfs
#| tbl-cap: "GTFS feeds used in the IMPT pipeline"

gtfs <- data.frame(
    Operator = c(
        "Carris Metropolitana", "Carris Municipal (Lisboa)", "MobiCascais",
        "Transportes Coletivos do Barreiro (TCB)",
        "Metropolitano de Lisboa (Metro)", "Comboios de Portugal (CP)",
        "Fertagus", "Metro Transportes do Sul (MTS)", "Transtejo/Soflusa (TTSL)"
    ),
    Mode = c("Bus", "Bus", "Bus", "Bus", "Metro", "Rail", "Rail", "Light Rail", "Ferry"),
    URL = c(
        "https://api.carrismetropolitana.pt/gtfs",
        "https://gateway.carris.pt/gateway/gtfs/api/v2.8/GTFS",
        "https://drive.google.com/u/0/uc?id=13ucYiAJRtu-gXsLa02qKJrGOgDjbnUWX&export=download",
        "https://backend.tcbarreiro.pt/download-gtfs",
        "https://metrolisboa.pt/google_transit/googleTransit.zip",
        "https://publico.cp.pt/gtfs/gtfs.zip",
        "https://mts.pt/imt/MTS-20240129.zip",
        "https://fertagus.pt/GTFSTMLzip/Fertagus_GTFS.zip",
        "https://api.transtejo.pt/files/GTFS.zip"
    )
)

# Convert URLs to clickable HTML links
gtfs$URL <- ifelse(grepl("^http", gtfs$URL),
    paste0("<a href=\"", gtfs$URL, "\" target=\"_blank\">", gtfs$URL, "</a>"),
    gtfs$URL
)

kbl(gtfs, col.names = c("Operator", "Mode", "GTFS URL"), booktabs = TRUE, escape = FALSE) |>
    kable_styling(font_size = 9, full_width = FALSE) |>
    column_spec(3, width = "8cm")
```

## R Packages {#sec-packages}

The following R packages are used across the pipeline:

```{r}
#| label: tbl-packages
#| tbl-cap: "Key R packages used in the IMPT pipeline"

pkgs <- data.frame(
    Package = c(
        "readr", "readxl", "openxlsx", "lubridate",
        "dplyr", "tidyr", "stringr", "tidyverse",
        "sf", "terra", "mapview", "stplanr", "h3jsr",
        "r5r", "accessibility", "odjitter", "tidytransit", "GTFShift",
        "FactoMineR", "assertthat", "piggyback"
    ),
    Purpose = c(
        "Fast CSV reading/writing", "Excel file reading (INE data)", "Excel file reading (safety data)", "Date/time manipulation",
        "Data manipulation", "Data reshaping", "String manipulation", "Collection of data science packages",
        "Spatial data handling (read, write, join, transform)", "DEM raster data processing", "Interactive map visualisation", "OD line-to-point conversion", "H3 hexagonal grid generation",
        "Multimodal travel time matrix computation", "Cumulative accessibility and cost-to-closest measures", "Origin-Destination jittering", "GTFS feed processing and stop frequencies", "GTFS loading, route frequency, filtering",
        "Principal Component Analysis", "Runtime assertion checks", "GitHub release asset upload/download"
    ),
    CRAN = c(
        "Yes", "Yes", "Yes", "Yes",
        "Yes", "Yes", "Yes", "Yes",
        "Yes", "Yes", "Yes", "Yes", "No*",
        "Yes", "Yes", "No*", "Yes", "No*",
        "Yes", "Yes", "Yes"
    )
)

kbl(pkgs,
    col.names = c("Package", "Primary purpose in IMPT", "CRAN?"),
    booktabs = TRUE, longtable = TRUE, escape = FALSE
) |>
    kable_styling(font_size = 9, full_width = FALSE) |>
    column_spec(2, width = "9cm") |>
    pack_rows("Data reading", 1, 4) |>
    pack_rows("Data manipulation", 5, 8) |>
    pack_rows("Geo spatial data", 9, 13) |>
    pack_rows("Routing and accessibility", 14, 18) |>
    pack_rows("PCA", 19, 19) |>
    pack_rows("Tests", 20, 20) |>
    pack_rows("Piggyback", 21, 21)
```

\* *Installed from GitHub: `odjitter` (`itsleeds/odjitter`, R subdirectory), `GTFShift` (`U-Shift/GTFShift`), `h3jsr` (`obrl-soil/h3jsr`).*
