---
title: "IMPT — Índice Multimodal de Pobreza nos Transportes"
subtitle: "Methodological Report"
author:
  - name: "U-Shift Lab, Instituto Superior Técnico"
    affiliation: "CERIS — Civil Engineering Research and Innovation for Sustainability"
date: today
date-format: "MMMM YYYY"
lang: en

format:
  html:
    toc: true
    toc-depth: 4
    toc-title: "Table of Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    highlight-style: github
    embed-resources: true
    fig-width: 8
    fig-height: 5
    include-in-header:
      text: |
        <style>
        body { font-family: 'Inter', sans-serif; }
        .callout { border-radius: 8px; }
        h1.title { font-size: 2.2em; color: #2c3e50; }
        h2 { color: #34495e; border-bottom: 2px solid #eee; padding-bottom: 4px; }
        h3 { color: #7f8c8d; }
        </style>
  pdf:
    toc: true
    toc-depth: 4
    number-sections: true
    documentclass: article
    geometry: margin=2.5cm
    fontsize: 11pt
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{xcolor}
        \usepackage{hyperref}
        \hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue}

bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Core packages used throughout this document for tables and figures
library(knitr)
library(kableExtra)
```

# Introduction {#sec-introduction}

This document describes the methodology behind the **Índice Multimodal de Pobreza nos Transportes** (IMPT) — a composite, multimodal transport poverty index developed for the **Lisbon Metropolitan Area** (Área Metropolitana de Lisboa, AML) by the U-Shift Lab at Instituto Superior Técnico.

The IMPT is designed to quantify the degree to which residents in different locations across the AML are disadvantaged in terms of access, mobility, affordability, and safety across multiple transport modes (car, public transit, cycling, walking). It is intended to be **reproducible and replicable** for other metropolitan areas in Portugal.

The index covers **`r "[FILL: total number of]"` parishes** (*freguesias*) and **18 municipalities** (*municípios*) in the Grande Lisboa and Península de Setúbal NUTS-II sub-regions, corresponding to approximately **2.87 million inhabitants** (Census 2021).

## Study Area {#sec-area}

The study area corresponds to the Lisbon Metropolitan Area (AML), defined as the combination of the NUTS-III sub-regions **Grande Lisboa** and **Península de Setúbal**, using administrative boundaries from **CAOP 2024.1** (Carta Administrativa Oficial de Portugal, Direção-Geral do Território).

The administrative geography used throughout the pipeline includes three nested levels:

- **H3 Hexagonal Grid, Resolution 8** (primary unit for routing calculations): ~3,686 hexagons, edge length ≈ 531 m, average area ≈ 0.74 km²  
- **Parishes** (*freguesias*): 225 units (post-2013 reform boundaries, updated to CAOP 2024)  
- **Municipalities** (*municípios*): 18 units

::: {.callout-note}
## Administrative boundary update
Several parishes changed between the 2016 IMOB travel survey and the CAOP 2024 boundaries. A conversion table and area-weighted redistribution procedure was applied to ensure consistency (see @sec-data-prep-boundaries).
:::

## Conceptual Framework {#sec-framework}

The IMPT is structured around **four dimensions**, assessed across **four transport modes**:

| Dimension | Modes assessed |
|-----------|---------------|
| **Accessibility** | Car, PT, Cycling, Walking |
| **Mobility** | Car, PT, Cycling, Walking |
| **Affordability** | Car, PT |
| **Safety** | Car, Cycling, Walking |

: IMPT dimensional structure {#tbl-framework}

Each dimension is computed as a composite of multiple indicators, normalised to a [1–100] scale, and then aggregated using **Principal Component Analysis (PCA)** to produce a single dimensional score. An overall IMPT score is derived as the geometric mean (or entropy-weighted product) of the four dimension scores.

---

# Data Sources {#sec-data-sources}

## Primary Data Sources

```{r}
#| label: tbl-datasources
#| tbl-cap: "Primary data sources used in the IMPT pipeline"

sources <- data.frame(
  Code = c("CAOP", "BGRI21", "IMOB", "GTFS", "OSM", "CM", "ANSR", "INE-Inc", "INE-Hous", "GBA", "COS", "IP"),
  Name = c(
    "CAOP 2024.1 — Administrative Boundaries",
    "Census 2021 — Base Geográfica de Referenciação de Informação",
    "IMOB 2017/18 — Inquérito à Mobilidade nas Áreas Metropolitanas",
    "GTFS — General Transit Feed Specification (all AML operators)",
    "OpenStreetMap (OSM) — Road network and POIs",
    "Carris Metropolitana — Health centres and schools datasets",
    "ANSR — Road accident data (Sinistralidade Rodoviária)",
    "INE — Household income by parish (Rendimento agregado 2023)",
    "INE — Housing costs by parish (Habitação custos 2021)",
    "GBA / TML — Building heights and shared mobility docks",
    "COS 2023 v1 (Copernicus S2) — Land use / built-up areas",
    "Infraestruturas de Portugal — Toll cost API"
  ),
  Year = c("2024", "2021", "2017/18", "2026-02-04", "2024-04", "2024", "2019–2023", "2023", "2021", "[FILL]", "2023", "2024"),
  URL = c(
    "geo2.dgterritorio.gov.pt/caop",
    "mapas.ine.pt/download/filesGPG/2021/nuts3/BGRI21_170.zip",
    "[FILL: source URL/repository]",
    "Multiple operators — see §3.1.4",
    "github.com/U-Shift/SiteSelection/releases/tag/0.1",
    "github.com/carrismetropolitana/datasets",
    "[FILL: ANSR data request source]",
    "[FILL: INE download URL]",
    "[FILL: INE download URL]",
    "[FILL: TML/GBA data source]",
    "[FILL: Copernicus or DGT source URL]",
    "portagens.infraestruturasdeportugal.pt"
  ),
  stringsAsFactors = FALSE
)

kable(sources, col.names = c("Code", "Dataset", "Reference period", "Source"),
      booktabs = TRUE, longtable = TRUE) |>
  kable_styling(font_size = 9, full_width = FALSE) |>
  column_spec(2, width = "9cm") |>
  column_spec(4, width = "5.5cm")
```

::: {.callout-warning}
## Items to confirm
Fields marked `[FILL]` should be replaced with the correct URL or reference. These correspond to data accessed from the internal server (`ushift@alfa`) and not publicly listed in the code.
:::

## GTFS Transit Operators {#sec-gtfs-operators}

Travel time matrices for public transit and stop-level frequency statistics were computed using GTFS feeds downloaded on **04/02/2026** from the following operators:

```{r}
#| label: tbl-gtfs
#| tbl-cap: "GTFS feeds used in the IMPT pipeline"

gtfs <- data.frame(
  Operator = c("Carris Metropolitana", "Carris Municipal (Lisboa)", "Comboios de Portugal (CP)",
               "MobiCascais", "Metropolitano de Lisboa (Metro)", "Transtejo/Soflusa (TTSL)",
               "Fertagus", "Transportes Colectivos do Barreiro (TCB)", "Metro Transportes do Sul (MTS)"),
  Mode = c("Bus", "Bus", "Rail", "Bus", "Metro", "Ferry", "Rail", "Bus", "Light Rail"),
  URL = c(
    "api.carrismetropolitana.pt/gtfs",
    "gateway.carris.pt/gateway/gtfs/api/v2.8/GTFS",
    "publico.cp.pt/gtfs/gtfs.zip",
    "drive.google.com/…",
    "metrolisboa.pt/google_transit/googleTransit.zip",
    "api.transtejo.pt/files/GTFS.zip",
    "fertagus.pt/GTFSTMLzip/Fertagus_GTFS.zip",
    "tcbarreiro.pt/…",
    "mts.pt/imt/MTS-20240129.zip"
  )
)
kable(gtfs, col.names = c("Operator", "Mode", "GTFS URL"), booktabs = TRUE) |>
  kable_styling(font_size = 9, full_width = FALSE) |>
  column_spec(3, width = "7cm")
```

## R Packages {#sec-packages}

The following R packages are used across the pipeline:

```{r}
#| label: tbl-packages
#| tbl-cap: "Key R packages used in the IMPT pipeline"

pkgs <- data.frame(
  Package = c("sf", "dplyr", "tidyr", "readr", "stringr", "tidyverse",
              "r5r", "accessibility", "odjitter", "tidytransit", "GTFShift",
              "h3jsr", "terra", "mapview", "stplanr",
              "FactoMineR", "assertthat", "piggyback", "lubridate", "readxl", "openxlsx"),
  Purpose = c(
    "Spatial data handling (read, write, join, transform)",
    "Data manipulation",
    "Data reshaping",
    "Fast CSV reading/writing",
    "String manipulation",
    "Collection of data science packages",
    "Multimodal travel time matrix computation",
    "Cumulative accessibility and cost-to-closest measures",
    "Origin-Destination jittering",
    "GTFS feed processing and stop frequencies",
    "GTFS loading, route frequency, filtering",
    "H3 hexagonal grid generation",
    "DEM raster data processing",
    "Interactive map visualisation",
    "OD line-to-point conversion",
    "Principal Component Analysis",
    "Runtime assertion checks",
    "GitHub release asset upload/download",
    "Date/time manipulation",
    "Excel file reading (INE data)",
    "Excel file reading (safety data)"
  ),
  CRAN = c(rep("Yes", 5), "Yes", "Yes", "Yes", "No*", "Yes", "No*",
            "No*", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes")
)
kable(pkgs, col.names = c("Package", "Primary purpose in IMPT", "CRAN?"),
      booktabs = TRUE, longtable = TRUE) |>
  kable_styling(font_size = 9, full_width = FALSE) |>
  column_spec(2, width = "9cm")
```

\* *Installed from GitHub: `odjitter` (`itsleeds/odjitter`, R subdirectory), `GTFShift` (`U-Shift/GTFShift`), `h3jsr` (`obrl-soil/h3jsr`).*

---

# Data Preparation {#sec-data-prep}

This phase corresponds to scripts **`00a`**, **`00b`**, **`01`**, **`02_*`** in the pipeline.

## Infrastructure and Helper Functions {#sec-helper}

**`00a_impt_data_handle.R`** defines two core helper functions:

- `impt_read(path)`: reads CSV, GeoPackage, GeoJSON, or RDS files. When running on the `ushift@alfa` server, it reads from the local `/data/IMPT/` directory; on any other machine, it fetches via HTTPS from `https://impt.server.ushift.pt` using an API key.
- `impt_write(content, path)`: writes to the same location logic (server vs. local `data/` folder).

These functions abstract away the data storage location, making the pipeline portable.

**`00b_data_load.R`** sets up globally-used variables (study area boundaries, grid, etc.) that are reused across scripts.

## Administrative Boundaries {#sec-data-prep-boundaries}

**Source:** CAOP 2024.1 (DGT — Direção-Geral do Território)

The full continental Portugal polygon file is filtered to the **Grande Lisboa** and **Península de Setúbal** NUTS-II sub-regions. For the municipality of Lisboa, parish boundaries with the Tagus estuary are replaced by clipped geometries (`FREGUESIASgeo.gpkg` from MQAT project) to remove the water area.

For parishes with multiple polygon features (due to enclave situations), only the polygon with the largest area is retained (`slice_max(area_ha)`).

### Conversion of 2016 to 2024 Parish Codes {#sec-dicofre}

The IMOB 2017 travel survey uses 2016 parish codes (*DICOFRE*). Several parishes were merged or split between 2016 and 2024. A manual conversion table was created after inspecting the differences in QGIS:

| Old DICOFRE (2016) | New DICOFRE(s) (2024) |
|--------------------|-----------------------|
| 151007 | 151008, 151009, 151010 |
| 111127 | 111134, 111135 |
| 111123 | 111129, 111131, 111132 |
| 111126 | 111130, 111133 |

: Parish code conversions from 2016 to 2024 {#tbl-dicofre}

Trips associated with split parishes were redistributed proportionally to the area of the new parishes. Conservation of total trip counts was validated using `assertthat`.

## H3 Hexagonal Grid {#sec-grid}

```{mermaid}
%%| label: fig-grid-overview
%%| fig-cap: "H3 grid generation and spatial assignment"
flowchart LR
    AML["AML boundary\n(CAOP 2024)"] --> H3["h3jsr::polygon_to_cells\nResolution 8"]
    H3 --> Cells["3,686 hexagons\n~531m edge, ~0.74 km² avg"]
    Cells --> Centroids["Grid centroids\n(for routing origins/destinations)"]
    Cells --> GridIndex["Grid index table\n(grid_id ↔ freg_id ↔ mun_id)"]
```

The H3 hexagonal grid is generated at **Resolution 8** using the `h3jsr` package [@Uber_H3_2018]. Resolution 8 provides a spatial unit with an average edge length of ~531 m and average area of ~0.74 km², which offers a good balance between spatial detail and computational tractability for routing-based accessibility measures.

The H3 grid covers the AML with approximately **3,686 active hexagons** (those intersecting the administrative boundaries). Each hexagon is assigned to a parish and municipality via a cross-reference table (`grid_nuts.csv`).

## Census Data Processing {#sec-census}

**Source:** INE Census 2021, BGRI 2021 at the base statistical unit level (BGRI — Base Geográfica de Referenciação de Informação), NUTS-3 Lisboa (code 170).

Census BGRI polygons are converted to **centroids** and spatially joined to the 2024 parish boundaries to assign the current parish code. The following variables are extracted:

| Variable | INE field | Description |
|----------|-----------|-------------|
| `population` | `N_INDIVIDUOS` | Total residents |
| `households` | `N_NUCLEOS_FAMILIARES` | Total family nuclei |
| `youth` | `N_INDIVIDUOS_0_14` | Residents aged 0–14 |
| `elderly` | `N_INDIVIDUOS_65_OU_MAIS` | Residents aged 65+ |
| `women` | `N_INDIVIDUOS_M` | Female residents |
| `buildings` | `N_EDIFICIOS_CLASSICOS` | Classic buildings |
| `buildings_pre1945` | `N_EDIFICIOS_CONSTR_ANTES_1945` | Buildings built before 1945 |

: Census variables used in IMPT {#tbl-census-vars}

For parish- and municipality-level aggregation, census data from 2016-era parishes are redistributed to 2024 parishes using the area-weighted conversion table described in @sec-dicofre.

## Dasymetric Grid Population Redistribution (COS 2023) {#sec-cos}

**Source:** COS 2023 v1 (Copernicus Sentinel-2), residential land use classes

To achieve a more accurate population distribution within the H3 grid, a **dasymetric mapping** approach is applied using the Carta de Ocupação do Solo (COS) 2023. Residential areas are extracted with the following classes and assigned weighting factors:

| COS 2023 classification | Weight |
|------------------------|--------|
| Contínuas predominantemente verticais | 3.0 |
| Contínuas predominantemente horizontais | 1.0 |
| Descontínuas | 0.6 |
| Descontínuas esparsas | 0.3 |

: COS 2023 residential classes and dasymetric weights {#tbl-cos-weights}

The procedure (`02_census_grid_with_cos.R`):

1. Census BGRI polygons are intersected with COS residential patches and H3 grid cells.
2. For each BGRI fragment, a weighted score is computed as `fragment_area × weight`.
3. Each BGRI's population is distributed proportionally to the weighted score of each fragment.
4. Results are summed per H3 cell.

This recovers approximately 99.4% of the total AML population (2,854,660 vs. 2,870,208 in the census), a significant improvement over simple centroid assignment.

## Points of Interest (POIs) {#sec-pois}

POIs are collected from three sources: OpenStreetMap, Carris Metropolitana open datasets, and GTFS feeds. The following categories are used:

```{r}
#| label: tbl-pois
#| tbl-cap: "POI categories, sources, and filters"

pois <- data.frame(
  Category = c("Healthcare (all)", "Hospitals", "Primary care centres",
                "Supermarkets/grocery", "Green spaces", "Recreation",
                "Schools (all)", "Primary schools", "Transit stops (all)",
                "Transit stops (bus)", "Transit stops (mass transit)", "Jobs"),
  Identifier = c("health", "health_hospital", "health_primary",
                  "groceries", "greenspaces", "recreation",
                  "schools", "schools_primary", "transit", "transit_bus", "transit_mass", "jobs"),
  Source = c("CM datasets", "CM datasets", "CM datasets",
              "OSM", "OSM", "OSM",
              "CM datasets", "CM datasets",
              "GTFS (all operators)", "GTFS (Carris, CM, TCB, Cascais próxima)", "GTFS (CP, Fertagus, Metro, MTS, TTSL)",
              "IMOB 2017 (jittered)"),
  Filter = c("All health centres", "name contains 'Hospital' or 'Centro de Medicina'",
              "name contains 'Centro de Saúde'",
              "type %in% c('supermarket','convenience')",
              "group=='leisure' & type %in% c('park','garden')",
              "type %in% c('library','theatre','cinema','restaurant','cafe','bar','pub','pitch','fitness_station','swimming_pool','sports_centre','fitness_center')",
              "All school types", "pre_school==1 | basic_1==1 | basic_2==1 | basic_3==1",
              "All stops from all operators",
              "Bus operators only",
              "Rail, metro, ferry operators",
              "Jittered OD destinations (see §3.7)")
)
kable(pois, col.names = c("Category", "Identifier", "Source", "Filter / Description"),
      booktabs = TRUE, longtable = TRUE) |>
  kable_styling(font_size = 9, full_width = FALSE) |>
  column_spec(4, width = "7cm")
```

OSM POI data are sourced from the `osm_poi_landuse.gpkg` file published in the [U-Shift SiteSelection repository](https://github.com/U-Shift/SiteSelection/releases/tag/0.1), representing OpenStreetMap data circa April 2024.

## Travel Survey Data and OD Jittering {#sec-jittering}

**Source:** IMOB 2017/18 — Inquérito à Mobilidade nas Áreas Metropolitanas do Porto e de Lisboa (INE / AMT). `[FILL: add official source URL/reference]`

The IMOB survey provides an OD matrix at parish level with modal split (walk, bike, car, public transit, other). Since accessibility and cost computations require disaggregated spatial locations, **jittering** is applied to generate point-level OD pairs.

### Jittering Algorithm

Jittering is performed using the `odjitter` R package [@Lovelace_odjitter_2022], following the approach described in @Lovelace_2022:

> *"OD pairs are sampled, and weights representing the importance of different trip 'originators' and 'attractors' are used to disaggregate trips."*

The road network (OSM) is used as the sub-point layer, so that jittered origin and destination points are located on the road network within each parish. A **disaggregation threshold of 200 trips** is applied, meaning OD pairs with more than 200 trips are split into multiple jittered pairs.

After jittering, each OD pair is assigned to a H3 grid cell for routing. The process generates approximately **18,997 jittered OD pairs** representing the AML commuting pattern.

## Road Network and DEM {#sec-network}

**Road network:** Exported from [HOT Export Tool](https://export.hotosm.org/exports/4782f0b8-6778-4c0e-8e4f-97fc62e7f240) in `.pbf` format for use with `r5r`. The network includes all OSM road types and GTFS transit feeds.

**Elevation:** Digital Elevation Model at 30 m resolution downloaded from [Copernicus Browser](https://browser.dataspace.copernicus.eu/) (Copernicus DEM GLO-30). Used to account for topographic effects on walking and cycling travel times via the Minetti metabolic-equivalent cost function.

The `r5r` network is built using `r5r::build_network()` with the `elevation = "MINETTI"` option.

## Modal Share from Census {#sec-modal-share}

**Source:** Census 2021 — question on mode used for the daily commute.

Script `02_census_modalShare.R` extracts and computes modal shares (private vehicle, public transit, active modes) at parish and grid level. These are used to compute **modal-share-weighted affordability** metrics (see @sec-affordability).

## Vehicle Ownership {#sec-veh-ownership}

**Source:** IMOB 2017 data (`02_veh_ownership.R`).

Vehicle ownership variables include average motorised vehicles per household and percentage of households without a vehicle, aggregated at parish and municipality level.

---

# Transport Routing — Travel Time Matrix {#sec-ttm}

**Script:** `03_ttm_gridh3.R`  
**Package:** `r5r` [@Pereira_r5r_2021]

A travel time matrix (TTM) is computed between all H3 grid centroids for the AML, for each transport mode, using `r5r::travel_time_matrix()`. The parameters used are:

```{r}
#| label: tbl-ttm-params
#| tbl-cap: "Travel time matrix parameters"

params <- data.frame(
  Parameter = c("Spatial unit", "Percentile", "Max trip duration", "Modes",
                "PT egress mode", "PT max walking time", "PT departure datetime",
                "PT max transfers (scenario 1)", "PT max transfers (scenario 2)",
                "Cycling max LTS", "Elevation model"),
  Value = c("H3 Resolution 8 centroids (~3,686 origins)", "50th percentile (median)",
            "120 minutes", "Walk, Cycling, Car, Transit (×2 transfer scenarios)",
            "WALK", "15 minutes",
            "Wednesday 04/02/2026 at 08:00 AM",
            "max_rides = 2 (1 transfer)", "max_rides = 3 (2 transfers)",
            "LTS ≤ 3 (comfortable streets)", "MINETTI metabolic-equivalent")
)
kable(params, booktabs = TRUE) |>
  kable_styling(font_size = 10, full_width = FALSE) |>
  column_spec(2, width = "10cm")
```

::: {.callout-note}
The 50th percentile travel time is used so that results represent typical (median) travel conditions rather than best- or worst-case scenarios. All parameters not listed here use `r5r` defaults.
:::

---

# Dimension Calculations {#sec-dimensions}

## Accessibility {#sec-accessibility}

**Script:** `04_access_opportunities.R`  
**Package:** `accessibility` [@Pereira_accessibility_2022]

```{mermaid}
%%| label: fig-accessibility-flow
%%| fig-cap: "Accessibility computation flowchart"
flowchart TB
    TTM["Travel Time Matrix\n5 modes × 2 transfer scenarios"] --> CumCut
    POIs["POIs per H3 cell\n(13 categories, spatially joined)"] --> CumCut
    Census["Population per H3 cell\n(residents, kids, elder, active)"] --> CumCut
    
    subgraph CumCut["Cumulative Accessibility"]
        direction TB
        Loop1["For each POI category"] --> Loop2["For each transport mode"]
        Loop2 --> Loop3["For each time cutoff\n5,10,15,30,45,60,75,90 min"]
        Loop3 --> Calc["accessibility::cumulative_cutoff()\nCount opportunities reachable\nwithin cutoff time"]
    end
    
    CumCut --> Agg["Aggregate to parish/municipality\n(weighted mean by population segment)"]
    Agg --> Out["Output: access_{poi}_{mode}_{time}_{census}"]
```

### Metric naming convention

Accessibility metrics follow the pattern:  
`access_{poi_type}_{mode}_{time}_{census_group}`

For example: `access_health_pt_30min_residents` = number of health facilities reachable by public transit within 30 minutes, for the total resident population.

### Aggregation to parish and municipality level

Grid-level accessibility values are aggregated to parishes and municipalities using a **population-weighted mean**, separately for each demographic group (residents, kids, active, elderly).

### Time to closest opportunity (mobility costs)

**Script:** `04_mobility_costs_time.R`  
**Function:** `accessibility::cost_to_closest()`

The travel time to the 1st, 2nd, and 3rd closest opportunity of each POI category is also computed, allowing measurement of baseline access time for the nearest facilities.

## Mobility {#sec-mobility}

Mobility indicators cover infrastructure availability and quality, public transport coverage and service levels, commuting patterns, and shared mobility provision.

### Walking infrastructure ratio

**Script:** `04_mobility.R`

The pedestrian infrastructure ratio is the proportion of road length that includes pedestrian infrastructure, computed using OSM data:

$$R_{ped} = \frac{\text{pedestrian path length}}{\text{road length}}$$

Road classification uses highway tags: `motorway`, `trunk`, `primary`, `secondary`, `tertiary`, `unclassified`, `residential`, `living_street`, plus link variants.

Pedestrian infrastructure uses tags: `highway = footway | pedestrian | steps | residential` and `footway = sidewalk | crossing | path | platform | corridor | alley | track`, and `sidewalk = both | left | right`.

### Cycling infrastructure ratio and quality

**Script:** `04_mobility.R`

The cycling infrastructure ratio is computed analogously to the pedestrian ratio. Cycling infrastructure data comes from the classified cycle network layer `AML_cycle_class.gpkg` `[FILL: source - TML/PMUS or classified from OSM?]`, which categorises cycleways into:

- Segregated cycleway (*ciclovia segregada*)
- Painted cycle lane (*ciclofaixa pintada na via*)
- Shared pedestrian infrastructure (*infraestrutura pedonal partilhada*)
- Shared road with vehicles, signed (*via partilhada com veículos motorizados, sinalizada*)

The cycling quality ratio is:
$$R_{cyc,q} = \frac{\text{segregated cycleway length}}{\text{total cycleway length}}$$

### PT stop coverage

**Script:** `04_mobility.R` / `04_mobility_transit.R`

The proportion of population within 500 m of a public transport stop is computed by:

1. Creating a 500 m buffer around every PT stop in the AML (all operators, from GTFS).
2. Counting BGRI centroids whose population falls within these buffers.
3. Dividing by total parish population.

**Limitation:** This is a binary coverage measure (served/not served); it does not account for service frequency or the number of modes available at each stop.

### PT headways and waiting times

**Script:** `04_mobility_transit.R`  
**Package:** `tidytransit`

Stop-level service frequencies are computed with `tidytransit::get_stop_frequency()` for four time windows:

| Period | Date | Time window |
|--------|------|-------------|
| Weekday peak | Wed 04/02/2026 | 08:00–09:00 |
| Weekday full day | Wed 04/02/2026 | 00:00–48:00 |
| Weekday night | Wed 04/02/2026 | 22:00–23:00 |
| Weekend | Sun 08/02/2026 | 10:00–11:00 |

: PT service frequency periods {#tbl-periods}

A 500 m buffer per stop is used to allocate stops to census population centroids. Results are aggregated to parishes and municipalities as **population-weighted mean headway and waiting time**.

Additionally, frequency reduction factors are computed to characterise night and weekend service reduction.

### Commuting travel times

**Script:** `04_mobility_commuting.R`

Using the jittered IMOB OD pairs and the travel time matrices, average commuting travel times are computed for each geographical unit and mode, weighted by number of trips. Trips with no feasible route within the constraints are assigned the maximum travel time (120 min) and counted separately as "not accomplishable" trips (PNA — *Percentagem Não Alcançáveis*).

```{mermaid}
%%| label: fig-commuting-flow
%%| fig-cap: "Commuting travel time computation"
flowchart TB
    TTM["Travel Time Matrix\n(all modes)"] --> Match
    ODs["Jittered OD pairs\n(IMOB 2017 commuting)"] --> Match
    Grid["H3 Grid"] --> Match
    
    subgraph Match["Spatial Assignment & Matching"]
        direction TB
        JoinGrid["Assign each OD\nto origin grid cell"] --> Lookup["Look up travel time\nfor each OD pair\nin TTM"]
        Lookup --> Fill["Fill missing:\nmax_duration = 120 min\n(NAs = not accomplishable)"]
    end
    
    Match --> AggMode["Aggregate by\ngeographic unit × mode"]
    AggMode --> Output["avg_tt_{mode}\ntrips, trips_na_{mode}\nPNA_{mode}"]
```

### Number of PT transfers

**Script:** `04_mobility_commuting.R` (line ~181)

For each commuting OD pair, the minimum number of transfers required is determined by checking whether a route exists in the matrices computed for 0, 1, 2, and 3 transfers. The minimum number of transfers for which a route exists is assigned to each OD pair, and population-weighted mean transfers are computed per geographical unit.

### Shared mobility

**Source:** TML (Transportes Metropolitanos de Lisboa) — shared mobility dock locations. `[FILL: URL or dataset reference]`

The number of shared mobility points (physical and virtual docks) is counted per parish and per H3 cell.

## Affordability {#sec-affordability}

**Script:** `04_affordability.R`, `04_mobility_costs_money_car.R`, `04_mobility_costs_money_pt.R`, `04_mobility_costs_money_pt_cost_structure.R`

Affordability is assessed by comparing transport and housing costs against household income. Commuting costs are computed for both car and public transport using `r5r::detailed_itineraries()` on the full set of jittered IMOB OD pairs, regardless of the reported travel mode.

### Car costs

**Script:** `04_mobility_costs_money_car.R`

Car commuting cost per trip = distance cost + tolls:

$$C_{car} = d \times 0.40\text{ €/km} + C_{toll}$$

The fixed cost of **0.40 €/km** is based on [Portaria n.º 1553-D/2008 de 31 de dezembro](https://diariodarepublica.pt/dr/detalhe/portaria/1553-d-2008-243733), which sets mileage allowance covering fuel and maintenance.

Tolls are computed via the [Infraestruturas de Portugal toll API](https://portagens.infraestruturasdeportugal.pt/) using the route geometry returned by `r5r::detailed_itineraries()`.

### Public transport costs — Two scenarios

**Scripts:** `04_mobility_costs_money_pt.R`, `04_mobility_costs_money_pt_cost_structure.R`

To account for uncertainty in PT fare payment practices, two cost scenarios are defined:

**Scenario A — Single fare (pre-Navegante baseline):**  
Based on Carris Lisboa revenue data:

- Total single ticket validations: 17,424,000
- Total single ticket revenue: 33,278,000 €
- Average cost per single ticket: **~1.91 €**

Free transfers within the same mode for 60 minutes are assumed. Different-mode transfers incur an additional charge.

**Scenario B — Monthly pass (Navegante):**  
Based on operator revenue data for monthly passes:

| Pass type | Volume (months) | Price |
|-----------|----------------|-------|
| Navegante Metropolitano | 5,063,510 | 40 € |
| Navegante Municipal | 887,190 | 30 € |
| Sub-23 | 1,103,782 | 20 € |
| Senior (65+) | 243,784 | 80 € |
| Free (youth) | 3,425,013 | 0 € |

Average monthly cost paid per pass holder: **~25.25 €**  
Assuming 44 journeys/month (2 trips × 22 working days): **~0.57 €/trip**

### Composite affordability indicators

Transport expenditure is expressed as a share of household income, with two variants (with and without housing costs):

$$\text{transp\_inc}_{car} = \frac{C_{car,year}}{I_{hh}}$$

$$\text{h\_transp\_inc}_{car} = \frac{C_{car,year} + C_{housing,year}}{I_{hh}}$$

A **modal-share-weighted composite** indicator combines car and PT costs proportionally to Census 2021 modal shares:

$$\text{transp\_inc\_comp} = \frac{C_{car} \cdot s_{car} + C_{PT} \cdot s_{PT}}{s_{car} + s_{PT} + s_{active}}$$

where $s_m$ is the modal share of mode $m$ from Census 2021.

Car costs are additionally adjusted for **vehicle occupancy rate** (from IMOB 2017, Table IV.7) and **household size** and **mobile population fraction** (from IMOB 2017, Table II.1), to convert per-trip costs to a household-level annual burden.

**Income data:** household income per parish from INE (*Rendimento agregado 2023*). `[FILL: confirm year and dataset reference]`  
**Housing costs:** median monthly housing cost per parish from INE (*Habitação custos 2021*). `[FILL: confirm dataset reference]`

::: {.callout-note}
## Canha parish
The parish of Canha (dtmnfr = 150701) is not served by any GTFS operator, resulting in NA PT costs. These values are imputed using the mean of neighbouring parishes within the same municipality (Montijo).
:::

## Safety {#sec-safety}

**Script:** `04_safety.R`  
**Source:** ANSR — Autoridade Nacional de Segurança Rodoviária, road accident georeferenced point data (2019–2023). `[FILL: exact data request reference or URL]`

```{mermaid}
%%| label: fig-safety-flow
%%| fig-cap: "Safety indicators computation"
flowchart TB
    Acc["ANSR Road Accidents\n2019–2023\n(georeferenced points)"] --> Filter
    Pop["Census Population\n(parish and grid)"] --> JoinPop
    Geo["Geographic units\n(parishes, municipalities, H3 grid)"] --> SJoin
    
    Filter["Filter: last 5 years\n(2019–2023)"] --> SJoin
    SJoin["st_join: assign accidents\nto geographic units"] --> ExcludeOuter["Exclude accidents\noutside built-up areas\n(ponts/bridges)"]
    ExcludeOuter --> Aggregate
    
    subgraph Aggregate["Aggregate per geographic unit"]
        direction TB
        TotalAcc["Total accidents (5 years)"] 
        Victims["Victim counts:\nfatal, serious, slight (30-day)"]
        Vehicles["Vehicles involved:\nmotorised, cycle, pedestrian"]
        Context["Context flags:\nat night, within localities"]
    end
    
    Aggregate --> JoinPop["Join census population"] --> CalcIdx
    
    subgraph CalcIdx["Compute indexes"]
        direction TB
        SevIdx["Severity index:\nfatalities / total victims"]
        ModeIdx["Mode-specific severity:\nfatalities / vehicles involved (per mode)"]
        Rate1000["Accident rate per 1,000 residents"]
        FatalRate["Fatality rate per 1,000 residents"]
    end
    
    CalcIdx --> BothSets["All accidents\n+\nWithin-localities only"]
```

### Accident data filtering

Accidents located outside administrative parish boundaries are excluded. This includes accidents on major bridges (Ponte 25 de Abril, Ponte Vasco da Gama) which cannot be assigned to a parish. Only accidents reported as occurring **within localities** (*dentro das localidades*) are used for the IMPT normalisation.

### Safety indicators computed

For each geographical unit and transport mode:

| Indicator | Formula |
|-----------|---------|
| Total accidents (5 years) | Count |
| Fatalities (30-day) | Sum `VM_30d` |
| Serious injuries (30-day) | Sum `FG_30d` |
| Slight injuries (30-day) | Sum `FL_30d` |
| Accident rate | `total_acidentes / population × 1000` |
| Fatality rate | `vitimas_mortais30 / population × 1000` |
| Severity index (total) | `fatalities / total_victims` |
| Severity index (motorised) | `fatalities / motorised_vehicles_involved` |
| Severity index (cycling) | `fatalities / cycling_involved` |
| Severity index (pedestrian) | `fatalities / pedestrians_involved` |

: Safety indicators {#tbl-safety-indicators}

---

# Index Computation — IMPT Calculator {#sec-calculator}

**Script:** `05_IMPTcalculator.R`  
**Package:** `FactoMineR` (PCA)

```{mermaid}
%%| label: fig-calculator-flow
%%| fig-cap: "IMPT index computation pipeline"
flowchart TB
    subgraph Inputs["Input Dimension Tables (by parish)"]
        AccIn["Accessibility indicators"]
        MobIn["Mobility indicators"]
        AffIn["Affordability indicators\n(two fare scenarios)"]
        SafIn["Safety indicators"]
    end
    
    Inputs --> Normalise
    
    subgraph Normalise["Normalisation to [1,100]"]
        direction TB
        Ben["Benefit indicators\nmin→1, max→100\nnormalize_benefit()"]
        Cost["Cost indicators\nmin→100, max→1\nnormalize_cost()"]
    end
    
    Normalise --> PCA
    
    subgraph PCA["PCA per dimension (global & per mode)"]
        direction TB
        Run["FactoMineR::PCA()\nextract PC1 scores"]
        Scale["Rescale PC1 to [1,100]\nscale_0_100()"]
        Direction["Validate direction:\ncheck correlation with\ntop indicator"]
    end
    
    PCA --> Composite
    
    subgraph Composite["Composite IMPT Score"]
        direction TB
        Geom["Geometric mean\n(Acc × Mob × Aff × Safety)^(1/4)"]
        EntrW["Entropy-weighted geometric mean\ncompute_entropy_weights()"]
        Simple["Simple arithmetic mean\n(Acc + Mob + Aff + Safety) / 4"]
    end
    
    Composite --> Agg["Population-weighted aggregation\nto municipality and grid levels"]
    Agg --> Out["Final IMPT scores:\nfreiguesia, municipality, grid\n× global and per mode"]
```

## Normalisation {#sec-normalisation}

All indicators are normalised to the range [1, 100] using min-max scaling:

**Benefit indicators** (higher raw value → better):
$$X_{norm} = 1 + \frac{X - X_{min}}{X_{max} - X_{min}} \times 99$$

**Cost indicators** (lower raw value → better):
$$X_{norm} = 1 + \frac{X_{max} - X}{X_{max} - X_{min}} \times 99$$

Accessibility metrics (`access_*`) are benefit indicators; mobility cost metrics (`mobility_cost_*`), PT waiting times, and all affordability indicators are cost indicators.

Missing values in accessibility indicators (arising from infeasible PT routes) are imputed with the mean of the respective variable prior to PCA.

## PCA-Based Dimension Scores {#sec-pca}

For each dimension (Accessibility, Mobility, Safety) and at the global (all-mode) and per-mode levels, a PCA is run using `FactoMineR::PCA()` with `dtmnfr` (parish ID) as a supplementary qualitative variable.

The **first principal component (PC1)** score is used as the dimension index, after rescaling to [1, 100]. PC1 captures the dominant axis of variation among the normalised indicators.

For **Affordability**, PCA is not applied (too few indicators); instead, the normalised composite `h_transp_inc_comp` (housing + transport cost as share of income, modal-share weighted) is used directly as the Affordability Index.

For **Safety per mode**, an equal-weight average of the severity index (inverted) and frequency (vehicle involvement count) is used instead of PCA.

## Composite IMPT Scores {#sec-composite}

Three aggregation methods are computed for the global IMPT and each per-mode IMPT:

1. **Geometric mean** (equal weights):
$$IMPT_{geom} = \left(\prod_{d} (S_d + 1)\right)^{1/n} - 1$$

2. **Arithmetic mean** (equal weights):
$$IMPT_{avg} = \frac{1}{n} \sum_d S_d$$

3. **Entropy-weighted geometric mean**:
$$IMPT_{entropy} = \prod_d (S_d + 1)^{w_d} - 1$$

where weights $w_d$ are derived from the **entropy weighting method**:

$$w_d = \frac{d_d}{\sum_j d_j}, \quad d_d = 1 - e_d, \quad e_d = -k \sum_i p_{id} \ln(p_{id})$$

with $k = 1/\ln(n)$ and $p_{id}$ being the proportion of the dimension score for area $i$.

## Spatial Aggregation {#sec-aggregation}

Once scores are computed at parish level, they are aggregated to higher levels:

- **Municipality level:** population-weighted mean of parish scores.
- **Grid level:** parishes are assigned to grid cells via the `grid_nuts.csv` cross-reference table; scores are assigned from the host parish (no further averaging at this stage).

---

# Output and Dashboard {#sec-output}

## Output Files

Final results are stored in three geographic formats for each spatial level:

| Level | Format | Variables |
|-------|--------|-----------|
| Parish (*freguesia*) | CSV + GeoPackage | All dimension scores + IMPT scores |
| Municipality (*município*) | CSV + GeoPackage | Population-weighted means |
| H3 Grid | CSV + GeoPackage | Assigned from host parish |

: Output file formats {#tbl-output}

Results are uploaded to the project data server at `https://impt.server.ushift.pt` using `piggyback` for publishing to the GitHub release.

## Dashboard Export {#sec-dashboard}

**Script:** `06b_dashboard_export.R`

The dashboard export script joins all indicator tables (accessibility, mobility, affordability, safety, census/land use, IMPT scores) with the spatial geometry files and exports them as three aggregated **GeoJSON** files for direct use in the web dashboard:

- `grid_aggregated.geojson`
- `freguesias_aggregated.geojson`
- `municipios_aggregated.geojson`

A champions lookup file (`champions.json`) is also exported, identifying the **top contributing indicators** to each dimension's PC1, for display in the dashboard sidebar.

## Web Dashboard

The IMPT results are visualised in an interactive web dashboard built with **React + TypeScript + Vite**, using **Leaflet/MapLibre** for map rendering. The dashboard is available at:

> `[FILL: public URL of the dashboard]`

The dashboard supports:

- Map visualisation by metric, mode, and geographic level
- Metric cards with legend, description, and top contributor indicators
- Dynamic IMPT score computation with user-defined dimension weights
- PT fare scenario switching (single fare vs. Navegante pass)

---

# Reproducibility {#sec-reproducibility}

## Code Repository

All data preparation and analysis scripts are available in the IMPT-data GitHub repository:

> `https://github.com/U-Shift/IMPT-data`

Scripts are organised with a numeric prefix indicating execution order:

| Prefix | Phase | Script(s) |
|--------|-------|-----------|
| 00 | Setup | `00a_impt_data_handle.R`, `00b_data_load.R` |
| 01 | Geometry & base data preparation | `01_data_prep.R` |
| 02 | Socio-demographics & land use | `02_*.R` (5 scripts) |
| 03 | Routing & travel time matrices | `03_ttm_gridh3.R` |
| 04 | Dimension calculations | `04_*.R` (10 scripts) |
| 05 | Index computation | `05_IMPTcalculator.R`, `05a_results_export_pca.R` |
| 06 | Output & dashboard | `06a_results_load.R`, `06b_dashboard_export.R` |

: Pipeline script organisation {#tbl-scripts}

## Replication for Other Areas

To replicate the IMPT for another Portuguese metropolitan area:

1. Replace the CAOP 2024 filter (NUTS-II codes `"Grande Lisboa"` and `"Península de Setúbal"`) with the target region.
2. Update the GTFS operator list in `01_data_prep.R`.
3. Obtain the IMOB travel survey OD matrix for the target area, and apply the dicofre conversion if needed.
4. Set up a HOT Export Tool bounding box for the new area.
5. Download the relevant Copernicus DEM tile.
6. Run scripts in order (00 → 06).

The H3 grid generation, routing engine, and scoring methodology are entirely generic and require no structural changes.

---

# References {#sec-references}

::: {#refs}
:::

---

# Appendix: PCA Diagnostics {#sec-appendix-pca}

The following table summarises the variance explained by PC1 in each PCA run:

```{r}
#| label: tbl-pca-summary
#| tbl-cap: "PCA diagnostics — variance explained by PC1 (indicative values; run 05_IMPTcalculator.R for up-to-date results)"

# This table would be populated from the PCA variance report CSV once available
# Placeholder structure:
pca_summary <- data.frame(
  Dimension_Mode = c("Accessibility (global)", "Mobility (global)", "Safety (global)",
                     "Accessibility – Car", "Accessibility – PT", "Accessibility – Bike", "Accessibility – Walk",
                     "Mobility – Car", "Mobility – PT", "Mobility – Bike", "Mobility – Walk",
                     "Safety – Car", "Safety – Bike", "Safety – Walk"),
  PC1_Variance = c("[run script]", "[run script]", "[run script]",
                   "[run script]", "[run script]", "[run script]", "[run script]",
                   "[run script]", "[run script]", "[run script]", "[run script]",
                   "[run script]", "[run script]", "[run script]"),
  Top_Variable = rep("[run script]", 14)
)

kable(pca_summary, col.names = c("Dimension / Mode", "PC1 Variance (%)", "Top contributing variable"),
      booktabs = TRUE) |>
  kable_styling(font_size = 10, full_width = FALSE)
```

::: {.callout-tip}
To populate this table automatically, source `05_IMPTcalculator.R` and read `PCA_Variance_Table.csv` and `PCA_Variance_and_Correlations.csv` from the results folder.
:::

---

*Report generated: `r format(Sys.time(), "%d %B %Y, %H:%M")`*  
*Pipeline version: see [CHANGELOG.md](https://github.com/U-Shift/IMPT-data/blob/main/CHANGELOG.md)*
