# Transport Routing — Travel Time Matrix {#sec-ttm}

**Script:** `03_ttm_gridh3.R`  
**Package:** `r5r` [@Pereira_r5r_2021]

```{r}
#| label: setup-ttm
#| include: false
library(knitr)
library(kableExtra)
```

A travel time matrix (TTM) is computed between all H3 grid centroids for the AML, for each transport mode, using `r5r::travel_time_matrix()`. The parameters used are:

```{r}
#| label: tbl-ttm-params
#| tbl-cap: "Travel time matrix parameters"

params <- data.frame(
    Parameter = c(
        "Spatial unit",
        "Percentile",
        "Max trip duration",
        "Modes",
        "PT egress mode",
        "PT max walking time",
        "PT departure datetime",
        "PT max transfers (scenario 1)",
        "PT max transfers (scenario 2)",
        "Cycling max Level of Traffic Stress",
        "Elevation model"
    ),
    Value = c(
        "H3 Resolution 8 centroids (~3,686 origins)",
        "50th percentile (median)",
        "120 minutes",
        "Walk, Bike, Car, Public Transit (×2 transfer scenarios)",
        "WALK",
        "15 minutes",
        "Wednesday 04/02/2026 at 08:00 AM",
        "max_rides = 2 (1 transfer)",
        "max_rides = 3 (2 transfers)",
        "LTS ≤ 3 (comfortable streets)",
        "MINETTI metabolic-equivalent"
    )
)
kable(params, booktabs = TRUE) |>
    kable_styling(font_size = 10, full_width = FALSE) |>
    column_spec(2, width = "10cm")
```

::: {.callout-note}
The 50th percentile travel time is used so that results represent typical (median) travel conditions rather than best- or worst-case scenarios. All parameters not listed here use `r5r` defaults.
:::

This process is long and computationally expensive. It is recommended to run it in a powerful machine.

The results are stored so we have a OD matrix of travel time between all H3 cells for each transport mode and scenario.
