---
title: "IMPT"
output: html_document
date: "2026-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# Load all necessary libraries here
library(dplyr)
library(readr)
library(ggplot2)
library(sf)        # If you are using spatial data for Freguesias
library(tidyr)
library(FactoMineR)
library(factoextra)
library(stringr)    # For string manipulation, if neededring)
library(janitor)     # For data cleaning, if needed
```



## The code to compute the IMPT


### Leaving out Mobility costs for now


```{r}
# Data loading Accessibility
freguesia_accessibility <- read_csv("/data/IMPT/results_data/freguesia_accessibility.csv")
Accessibility <- freguesia_accessibility |> select (-nuts)
```

```{r}
# Data loading Mobility
freguesias_infrastructure_ratio <- read_csv("/data/IMPT/mobility/freguesias_infrastructure_ratio.csv")
# names(freguesias_infrastructure_ratio)
freguesias_infrastructure_ratio = freguesias_infrastructure_ratio |> 
  select(dtmnfr, road_length, pedpath_length, cycleway_length, 
        segregated_cycleway_length, pedpath_to_road_ratio, cycleway_to_road_ratio, 
        cycling_quality_ratio)

freguesias_stops_coverage <- read_csv("/data/IMPT/mobility/freguesias_stops_coverage.csv")
# names(freguesias_stops_coverage)
#### There is not a varaible named "Stops_coverage", and the on at mauricio drive has different values than  the suspected one
# summary(Mobility2$Stops_coverage)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#  0.1337  0.7302  0.9211  0.8522  0.9833  1.0000 
# summary(freguesias_stops_coverage$ratio_served_population)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#  0.3500  0.9000  0.9800  0.9354  1.0000  1.0000 
 
freguesias_stops_coverage = freguesias_stops_coverage |> 
  select(freguesia, ratio_served_population) # will use this for now...
# we need to add the dtmnfr code to be able to join
freguesias_names = st_read("/data/IMPT/geo/freguesias_2024_unique.gpkg") |> 
  st_drop_geometry() |> 
  select(dtmnfr, freguesia)
freguesias_stops_coverage = freguesias_stops_coverage |>
  left_join(freguesias_names, by = "freguesia") |> 
  select(-freguesia) |> 
  mutate(dtmnfr = as.numeric(dtmnfr))

freguesias_shared_mobility <- read_csv("/data/IMPT/mobility/freguesias_shared_mobility.csv")
# names(freguesias_shared_mobility)
freguesias_shared_mobility = freguesias_shared_mobility |> 
  select(dtmnfr, shared_mobility_points)

freguesias_mobility <- read_csv("/data/IMPT/results_data/freguesia_mobility.csv")
names(freguesias_mobility)
freguesias_mobility = freguesias_mobility |> 
  select(-nuts)

freguesias_headways = read_csv("/data/IMPT/mobility_transit/freguesias_headways.csv") 
names(freguesias_headways)
  # summary(Mobility2$PT_Waitingtime)
  # summary(freguesias_headways$weighted_waiting_time_peak)
  # summary(Mobility2$PT_Weight_frequendy_peak_day)
  # summary(freguesias_headways$weighted_frequency_day)
freguesias_headways = freguesias_headways |> 
  select(dtmnfr, weighted_frequency_day, weighted_waiting_time_peak)

## Join all tables
Mobility <- freguesias_infrastructure_ratio |> 
  left_join(freguesias_stops_coverage, by = "dtmnfr") |> 
  left_join(freguesias_shared_mobility, by = "dtmnfr") |> 
  left_join(freguesias_mobility, by = "dtmnfr") |> 
  left_join(freguesias_headways, by = "dtmnfr")

# names(Mobility)
# names(Mobility2)
```




######### Data prepatation #########################

```{r}



# Affordability_post_nav <- freguesia_affordability |>
#   mutate(h_transp_inc_car = (housing + affordability_car_total_money*365)/Income) |> 
#   mutate(h_transp_inc_pt = (housing + affordability_transit_pass_total_money*365)/Income) |>
#   mutate(transp_inc_car = (affordability_car_total_money*365)/Income) |>
#   mutate(transp_inc_pt = (affordability_transit_pass_total_money*365)/Income) |>
#   select(dtmnfr, h_transp_inc_car, h_transp_inc_pt, transp_inc_car, transp_inc_pt)
# 
# #Replace NA with 0 for affordability dataset (if any)
# Affordability_post_nav <- Affordability_post_nav |> 
#   mutate(across(where(is.numeric), ~ ifelse(is.na(.x), 0, .x)))
# 
# mean_viz_canha <- Affordability_post_nav |> # Filter dtmnfr starting with 1507
#   filter(str_starts(dtmnfr, "1507")) |> 
#   filter(dtmnfr!= "150701")|>
#   summarise(mean_h_transp_inc_pt = mean(h_transp_inc_pt, na.rm = TRUE),
#             mean_transp_inc_pt = mean(transp_inc_pt, na.rm = TRUE))
# 
# 
# #### Replace zeros with the municipality avg
#   
# Affordability_post_nav <- Affordability_post_nav |> 
#   mutate(h_transp_inc_pt = case_when(
#      dtmnfr == "150701" ~ mean_viz_canha$mean_h_transp_inc_pt,
#     TRUE ~ h_transp_inc_pt
#   )) |>
#   mutate(transp_inc_pt = case_when(
#      dtmnfr == "150701" ~ mean_viz_canha$mean_transp_inc_pt,
#     TRUE ~ transp_inc_pt))
#     
# 
# 
#     
#     
# Affordability_pre_nav <- freguesia_affordability |> 
#   mutate(h_transp_inc_car = (housing + affordability_car_total_money*365)/Income) |>
#   mutate(h_transp_inc_pt = (housing + affordability_transit_single_fare_total_money*365)/Income) |>
#   mutate(transp_inc_car = (affordability_car_total_money*365)/Income) |>
#   mutate(transp_inc_pt = (affordability_transit_single_fare_total_money*365)/Income) |>
#   select(dtmnfr, h_transp_inc_car, h_transp_inc_pt, transp_inc_car, transp_inc_pt)
# 
# #Replace NA with 0 for affordability dataset (if any)
# Affordability_pre_nav <- Affordability_pre_nav |> 
#   mutate(across(where(is.numeric), ~ ifelse(is.na(.x), 0, .x)))
#   
# mean_viz_canha2 <- Affordability_pre_nav |> # Filter dtmnfr starting with 1507
#   filter(str_starts(dtmnfr, "1507")) |> 
#   filter(dtmnfr!= "150701")|>
#   summarise(mean_h_transp_inc_pt = mean(h_transp_inc_pt, na.rm = TRUE),
#             mean_transp_inc_pt = mean(transp_inc_pt, na.rm = TRUE))
# 
# 
# #### Replace zeros with the municipality avg
#   
# Affordability_pre_nav <- Affordability_pre_nav |> 
#   mutate(h_transp_inc_pt = case_when(
#      dtmnfr == "150701" ~ mean_viz_canha2$mean_h_transp_inc_pt,
#     TRUE ~ h_transp_inc_pt
#   )) |>
#   mutate(transp_inc_pt = case_when(
#      dtmnfr == "150701" ~ mean_viz_canha2$mean_transp_inc_pt,
#     TRUE ~ transp_inc_pt))
# 
# Safety <- accidents_by_freguesia_5years_dentrolocalidades|> 
#   select(-2, -3, -4)
# 
# # Replace NA with 0 for safety dataset
# Safety <- Safety |> 
#   mutate(across(where(is.numeric), ~ ifelse(is.na(.x), 0, .x)))
# 
#   
  
  
  
```


######## Chunk for specific mode dimensions datasets


```{r}

###################### Create a dataset per mode per dimension #####################

Accessibility_car <- Accessibility |> 
  select(dtmnfr, contains("_car_"))
Accessibility_pt <- Accessibility |>
  select(dtmnfr, contains("_transit_"))
Accessibility_bike <- Accessibility |>
  select(dtmnfr, contains("_bike_"))
Accessibility_walk <- Accessibility |>
  select(dtmnfr, contains("_walk_"))


Mobility_car <- Mobility |> 
  select(dtmnfr, contains("_car"),"road_length")
Mobility_pt <- Mobility |>
  select(dtmnfr, contains("PT"), "Stops_coverage", contains("transit"), Shared_mobility)
Mobility_bike <- Mobility |>
  select(dtmnfr, contains("cycl"), contains("bike"))
Mobility_walk <- Mobility |>
  select(dtmnfr, contains("_walk"),contains("ped"))

Affordability_car <- Affordability_post_nav |> 
  select(dtmnfr, contains("car"))
Affordability_pt_post_nav <- Affordability_post_nav |>
  select(dtmnfr, contains("pt"))
Affordability_pt_pre_nav <- Affordability_pre_nav |>
  select(dtmnfr, contains("pt"))


  
Safety_car <- Safety |> 
  select(dtmnfr, contains("car"), contains("motorizado"))
Safety_bike <- Safety |>
  select(dtmnfr, contains("bici"))

#Replace NA with 0 for bike safety

Safety_bike <- Safety_bike |> 
  mutate(across(where(is.numeric), ~ ifelse(is.na(.x), 0, .x)))


Safety_walk <- Safety |>
  select(dtmnfr, contains("peo"))


```




### Step 1: Normalize the indicators

Use Normalize benefit for indicators where higher is better and Normalize cost for indicators where lower is better.
```{r cars}

# Use for "More is Better" (e.g., Number of stops, Frequency, Safety score)
normalize_benefit <- function(x) {
  x_min <- min(x, na.rm = TRUE)
  x_max <- max(x, na.rm = TRUE)
  
  # Standard Min-Max scaled to 1-100
  return( 1 + ( (x - x_min) / (x_max - x_min) ) * (100 - 1) )
}

# Use for "More is Worse" (e.g., Travel time, Cost, Distance to station)
normalize_cost <- function(x) {
  x_min <- min(x, na.rm = TRUE)
  x_max <- max(x, na.rm = TRUE)
  
  # Inverted Min-Max scaled to 1-100
  # This makes the lowest raw value become 100
  return( 1 + ( (x_max - x) / (x_max - x_min) ) * (100 - 1) )
}


Accessibility_Norm <- Accessibility |> 
  mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with("access"), ~ normalize_benefit(.x)),
    
    # Same for your negative indicators
    across(starts_with("mobility"), ~ normalize_cost(.x))
  ) 


Mobility_Norm <- Mobility|> 
  mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with(c("PT_Wait", "mobility")), ~ normalize_cost(.x)),
    
    # Same for the negative indicators
    across(!c(dtmnfr, starts_with(c("PT_Wait", "mobility"))), ~ normalize_benefit(.x))
  )


Affordability_Norm_post_nav <- Affordability_post_nav |> ### All are negative
  mutate(
    
    across(!dtmnfr, ~ normalize_cost(.x)))

Affordability_Norm_pre_nav <- Affordability_pre_nav |> ### All are negative
  mutate(
    
    across(!dtmnfr, ~ normalize_cost(.x)))


Safety_Norm <- Safety |> ### All are negative
    mutate(
    across(!dtmnfr, ~ normalize_cost(.x)))
    
##### Save normalized dataset

write_csv(Mobility_Norm, "Mobility_Norm.csv")


#replace NA with the mean
Accessibility_Norm <- Accessibility_Norm |> 
  mutate(
    across(where(is.numeric), ~ ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x))
  )
write_csv(Accessibility_Norm, "Accessibility_Norm.csv")
write_csv(Affordability_Norm_post_nav, "Affordability_Norm_post_nav.csv")
write_csv(Affordability_Norm_pre_nav, "Affordability_Norm_pre_nav.csv")
write_csv(Safety_Norm, "Safety_Norm.csv")
  
```

### Normalize indicators for per mode datasets

The functions to normalize are in the previous chunk

```{r}



Accessibility_car_Norm <- Accessibility_car |> 
  mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with("access"), ~ normalize_benefit(.x)),
    
    # Same for your negative indicators
    across(starts_with("mobility"), ~ normalize_cost(.x))
  ) 

Accessibility_pt_Norm <- Accessibility_pt |> 
  mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with("access"), ~ normalize_benefit(.x)),
    
    # Same for your negative indicators
    across(starts_with("mobility"), ~ normalize_cost(.x))
  )

Accessibility_bike_Norm <- Accessibility_bike |> 
  mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with("access"), ~ normalize_benefit(.x)),
    
    # Same for your negative indicators
    across(starts_with("mobility"), ~ normalize_cost(.x))
  )

Accessibility_walk_Norm <- Accessibility_walk |>
  mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with("access"), ~ normalize_benefit(.x)),
    
    # Same for your negative indicators
    across(starts_with("mobility"), ~ normalize_cost(.x))
  )




Mobility_car_Norm <- Mobility_car|> 
  mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with(c("mobility")), ~ normalize_cost(.x)),
    
    # Same for the negative indicators
    across(!c(dtmnfr, starts_with(c("mobility"))), ~ normalize_benefit(.x))
  )
Mobility_pt_Norm <- Mobility_pt|>
 mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with(c("PT_Wait", "mobility")), ~ normalize_cost(.x)),
    
    # Same for the negative indicators
    across(!c(dtmnfr, starts_with(c("PT_Wait", "mobility"))), ~ normalize_benefit(.x))
  )
                                                   
Mobility_bike_Norm <- Mobility_bike|>
 mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with(c("mobility")), ~ normalize_cost(.x)),
    
    # Same for the negative indicators
    across(!c(dtmnfr, starts_with(c("mobility"))), ~ normalize_benefit(.x))
  )

Mobility_walk_Norm <- Mobility_walk|>
   mutate(
    # Use c() inside starts_with for multiple prefixes
    across(starts_with(c("mobility")), ~ normalize_cost(.x)),
    
    # Same for the negative indicators
    across(!c(dtmnfr, starts_with(c("mobility"))), ~ normalize_benefit(.x))
  )



Affordability_car_Norm <- Affordability_car |> ### All are negative
  mutate(
    
    across(!dtmnfr, ~ normalize_cost(.x)))

Affordability_pt_Norm_post_nav <- Affordability_pt_post_nav |> ### All are negative
  mutate(
    
    across(!dtmnfr, ~ normalize_cost(.x)))

Affordability_pt_Norm_pre_nav <- Affordability_pt_pre_nav |> ### All are negative
  mutate(
    
    across(!dtmnfr, ~ normalize_cost(.x)))



Safety_car_Norm <- Safety_car |> ### All are negative
  mutate(
    
    across(!dtmnfr, ~ normalize_cost(.x)))

Safety_bike_Norm <- Safety_bike |> ### All are negative
  mutate(
    
    across(!dtmnfr, ~ normalize_cost(.x)))
Safety_walk_Norm <- Safety_walk |> ### All are negative
  mutate(
    
    across(!dtmnfr, ~ normalize_cost(.x)))



##### Save normalized dataset

write_csv(Mobility_car_Norm, "Mobility_car_Norm.csv")
write_csv(Mobility_pt_Norm, "Mobility_pt_Norm.csv")
write_csv(Mobility_bike_Norm, "Mobility_bike_Norm.csv")
write_csv(Mobility_walk_Norm, "Mobility_walk_Norm.csv")

#replace NA with the mean
Accessibility_pt_Norm <- Accessibility_pt_Norm |> 
  mutate(
    across(where(is.numeric), ~ ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x))
  )

write_csv(Accessibility_car_Norm, "Accessibility_car_Norm.csv")
write_csv(Accessibility_pt_Norm, "Accessibility_pt_Norm.csv")
write_csv(Accessibility_bike_Norm, "Accessibility_bike_Norm.csv")
write_csv(Accessibility_walk_Norm, "Accessibility_walk_Norm.csv")

write_csv(Affordability_car_Norm, "Affordability_car_Norm.csv")
write_csv(Affordability_pt_Norm_pre_nav, "Affordability_pt_Norm_pre_nav.csv")
write_csv(Affordability_pt_Norm_post_nav, "Affordability_pt_Norm_post_nav.csv")


write_csv(Safety_car_Norm, "Safety_car_Norm.csv")
write_csv(Safety_bike_Norm, "Safety_bike_Norm.csv")
write_csv(Safety_walk_Norm, "Safety_walk_Norm.csv")


```




## Step 2: Compute dimensions weights based on indicators

We will use two approaches:

- Equal weights: All indicators contribute equally to the dimension score.

```{r pressure, echo=FALSE}

# Load necessary libraries


# Equal weights
# Function to calculate the Dimension Score (Equal Weights for Indicators)
# calculate_dimension_EW <- function(df_norm) {
#   # We exclude the ID column (dtmnfr) and take the average of all numeric columns
#   score <- rowMeans(df_norm[, -1], na.rm = TRUE)
#   
# }
# 
# # Calculate the 4 dimensions
# Acc_EW_Score  <- calculate_dimension_EW(Accessibility_Norm)
# Aff_EW_Score  <- calculate_dimension_EW(Affordability_Norm)
# Mob_EW_Score  <- calculate_dimension_EW(Mobility_Norm)
# Saf_EW_Score  <- calculate_dimension_EW(Safety_Norm)
# 
# # Combine them into a new Results table
# Results_Equal_Weights <- data.frame(
#   dtmnfr = Accessibility_Norm$dtmnfr,
#   Accessibility = Acc_EW_Score,
#   Affordability = Aff_EW_Score,
#   Mobility      = Mob_EW_Score,
#   Safety        = Saf_EW_Score
# )


```


#### Equal weights per mode

```{r,eval=FALSE}

# Calculate the 4 dimensions
Acc_car_EW_Score  <- calculate_dimension_EW(Accessibility_car_Norm)
Acc_pt_EW_Score  <- calculate_dimension_EW(Accessibility_pt_Norm)
Acc_bike_EW_Score  <- calculate_dimension_EW(Accessibility_bike_Norm)
Acc_walk_EW_Score  <- calculate_dimension_EW(Accessibility_walk_Norm)


Aff_car_EW_Score  <- calculate_dimension_EW(Affordability_car_Norm)
Aff_pt_EW_Score  <- calculate_dimension_EW(Affordability_pt_Norm)
Aff_bike_EW_Score  <- calculate_dimension_EW(Affordability_bike_Norm)


Mob_car_EW_Score  <- calculate_dimension_EW(Mobility_car_Norm)
Mob_pt_EW_Score  <- calculate_dimension_EW(Mobility_pt_Norm)
Mob_bike_EW_Score  <- calculate_dimension_EW(Mobility_bike_Norm)
Mob_walk_EW_Score  <- calculate_dimension_EW(Mobility_walk_Norm)

Saf_car_EW_Score  <- calculate_dimension_EW(Safety_car_Norm)
Saf_bike_EW_Score  <- calculate_dimension_EW(Safety_bike_Norm)
Saf_walk_EW_Score  <- calculate_dimension_EW(Safety_walk_Norm)

# Combine them into a new Results table
Results_Equal_Weights_car <- data.frame(
  dtmnfr = Accessibility_car_Norm$dtmnfr,
  Accessibility = Acc_car_EW_Score,
  Affordability = Aff_car_EW_Score,
  Mobility      = Mob_car_EW_Score,
  Safety        = Saf_car_EW_Score
)

Results_Equal_Weights_pt <- data.frame(
  dtmnfr = Accessibility_pt_Norm$dtmnfr,
  Accessibility = Acc_pt_EW_Score,
  Affordability = Aff_pt_EW_Score,
  Mobility      = Mob_pt_EW_Score,
  Safety        = NA # No safety indicators for PT
)

Results_Equal_Weights_bike <- data.frame(
  dtmnfr = Accessibility_bike_Norm$dtmnfr,
  Accessibility = Acc_bike_EW_Score,
  Affordability = Aff_bike_EW_Score,
  Mobility      = Mob_bike_EW_Score,
  Safety        = Saf_bike_EW_Score
)

Results_Equal_Weights_walk <- data.frame(
  dtmnfr = Accessibility_walk_Norm$dtmnfr,
  Accessibility = Acc_walk_EW_Score,
  Affordability = NA, # No affordability indicators for walking
  Mobility      = Mob_walk_EW_Score,
  Safety        = Saf_walk_EW_Score
)
```






### Step 3: Compute USING PCA: We will perform Principal Component Analysis (PCA) to determine the weights based on the variance explained by each indicator. Using PC1.
```{r}


##################### PCA #####################################


# Perform PCA on the normalized indicators
# Assuming 'data' is your dataset with normalized indicators
pca_access <- PCA(Accessibility_Norm, quali.sup = 1, graph = FALSE)
#pca_afford_post_nav <- PCA(Affordability_Norm_post_nav, quali.sup = 1, graph = FALSE)
#pca_afford_pre_nav <- PCA(Affordability_Norm_pre_nav, quali.sup = 1, graph = FALSE)
pca_mobility <- PCA(Mobility_Norm, quali.sup = 1, graph = FALSE) 
pca_safety <- PCA(Safety_Norm, quali.sup = 1, graph = FALSE)



################ AFFORDABILITY NOT PCA #######################################
afford_pre_nav <- Affordability_Norm_pre_nav |> 
  left_join(census_modal_share, by = "dtmnfr")

afford_post_nav <- Affordability_Norm_post_nav |> 
left_join(census_modal_share, by = "dtmnfr") |>  
  mutate(MS_CAR_AFF = h_transp_inc_car * private_vehicle_share) |>
  mutate(MS_PT_AFF = h_transp_inc_pt * pt_share) |>
  mutate(MS_AFF_POST_NAV = (MS_CAR_AFF + MS_PT_AFF) / (private_vehicle_share + pt_share + active_share))

afford_pre_nav <- Affordability_Norm_pre_nav |> 
  left_join(census_modal_share, by = "dtmnfr") |> 
  mutate(MS_CAR_AFF = h_transp_inc_car * private_vehicle_share) |>
  mutate(MS_PT_AFF = h_transp_inc_pt * pt_share) |> 
  mutate(MS_AFF_PRE_NAV = (MS_CAR_AFF + MS_PT_AFF) / (private_vehicle_share + pt_share + active_share))




afford_post_nav_fin <- afford_post_nav |> 
  select(dtmnfr, MS_AFF_POST_NAV)


afford_pre_nav_fin <- afford_pre_nav |> 
  select(dtmnfr, MS_AFF_PRE_NAV)

# A helper function to scale any vector 0-100
scale_0_100 <- function(x) {
  x_min <- min(x, na.rm = TRUE)
  x_max <- max(x, na.rm = TRUE)
  
  # Logic: 1 + (normalized_value * 99) 
  # This ensures the lowest value is 1 and the highest is 100
  1 + ( (x - x_min) / (x_max - x_min) * 99 )
}

# 1. Extract Accessibility Score
acc_scores <- data.frame(
  dtmnfr = Accessibility_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(pca_access$ind$coord[, 1])
)

# 2. Extract Affordability Score
#aff_scores_post_nav <- data.frame(
  #dtmnfr = Affordability_Norm_post_nav$dtmnfr,
  #Affordability_Index_post_nav = scale_0_100(pca_afford_post_nav$ind$coord[, 1])
#)

#aff_scores_pre_nav <- data.frame(
 # dtmnfr = Affordability_Norm_pre_nav$dtmnfr,
  #Affordability_Index_pre_nav = scale_0_100(pca_afford_pre_nav$ind$coord[, 1])
#)

# 3. Extract Safety Score
saf_scores <- data.frame(
 dtmnfr = Safety_Norm$dtmnfr,
  Safety_Index = scale_0_100(pca_safety$ind$coord[, 1])
)

#4. Extract Mobility Score
mob_scores <- data.frame(
  dtmnfr = Mobility_Norm$dtmnfr,
  Mobility_Index = scale_0_100(pca_mobility$ind$coord[,1])
)


#### Join Results scores

Final_Results_Master <- acc_scores |> 
  #left_join(aff_scores_pre_nav, by = "dtmnfr") |> 
  #left_join(aff_scores_post_nav, by = "dtmnfr") |>
  left_join(afford_pre_nav_fin, by = "dtmnfr") |>
   left_join(afford_post_nav_fin, by = "dtmnfr") |>
  left_join(mob_scores, by = "dtmnfr") |>
  left_join(saf_scores, by = "dtmnfr")

#Extract the weights from the first principal component 


#pca_weights_access <- pca_access$var$coord[, 1] / sum(pca_access$var$coord[, 1])
#pca_weights_afford <- pca_afford$var$coord[, 1] / sum(pca_afford$var$coord[, 1])
#pca_weights_mobility <- pca_mobility$var$coord[, 1] / sum(pca_mobility$var$coord[, 1])
#pca_weights_safety <- pca_safety$var$coord[, 1] / sum(pca_safety$var$coord[, 1])



####CHECK THE CHAMPIONS INDICATORS

Champions <- function(pca_here, dim = 1, n = 10) {
  # Extract the specific dimension contributions
  contributions <- pca_here$var$contrib[, dim]
  
  # Sort and take the top n
  # Remember: use TRUE (all caps)
  top_n <- sort(contributions, decreasing = TRUE) |> head(n)
  
  return(top_n)
}

Champions_access <- Champions(pca_access)
#Champions_afford_pre_nav <- Champions(pca_afford_pre_nav) # or pca_afford_post_nav, depending on which you want to analyze
#Champions_afford_post_nav <- Champions(pca_afford_post_nav)
Champions_mobility <- Champions(pca_mobility)
Champions_safety <- Champions(pca_safety)




# Write the PCA scores to CSV files

write_csv(acc_scores, "Accessibility_PCA_Scores.csv")
#write_csv(aff_scores_pre_nav, "Affordability_PCA_Scores_pre_nav.csv")
#write_csv(aff_scores_post_nav, "Affordability_PCA_Scores_post_nav.csv")
write_csv(saf_scores, "Safety_PCA_Scores.csv")
write_csv(mob_scores, "Mobility_PCA_Scores.csv")

# Write champions scores
write_csv(data.frame(Indicator = names(Champions_access), Contribution = Champions_access), "Champions_Accessibility.csv")
#write_csv(data.frame(Indicator = names(Champions_afford_pre_nav), Contribution = Champions_afford_pre_nav), "Champions_Affordability_pre_nav.csv")
#write_csv(data.frame(Indicator = names(Champions_afford_post_nav), Contribution = Champions_afford_post_nav), "Champions_Affordability_post_nav.csv")
write_csv(data.frame(Indicator = names(Champions_mobility), Contribution = Champions_mobility), "Champions_Mobility.csv")
write_csv(data.frame(Indicator = names(Champions_safety), Contribution = Champions_safety), "Champions_Safety.csv")

                                                                           
```


#### PCA computations per mode

The function to rescale is on the previous chunk

```{r}

##################### PCA #####################################


# Perform PCA on the normalized indicators

pca_car_access <- PCA(Accessibility_car_Norm, quali.sup = 1, graph = FALSE)
pca_pt_access <- PCA(Accessibility_pt_Norm, quali.sup = 1, graph = FALSE)
pca_bike_access <- PCA(Accessibility_bike_Norm, quali.sup = 1, graph = FALSE)
pca_walk_access <- PCA(Accessibility_walk_Norm, quali.sup = 1, graph = FALSE)



#pca_car_afford <- PCA(Affordability_car_Norm, quali.sup = 1, graph = FALSE).  NOT ENOUGH INDICATORS
#pca_pt_afford_pre_nav <- PCA(Affordability_pt_Norm_pre_nav, quali.sup = 1, graph = FALSE) NOT ENOUGH INDICATORS
#pca_pt_afford_post_nav <- PCA(Affordability_pt_Norm_post_nav, quali.sup = 1, graph = FALSE) NOT ENOUGH INDICATORS

car_afford <- Affordability_car_Norm |> 
  select(dtmnfr, 2) 

pt_afford_pre_nav <- Affordability_pt_Norm_pre_nav |> 
  select(dtmnfr, 2)

pt_afford_post_nav <- Affordability_pt_Norm_post_nav |> 
  select(dtmnfr, 2)


pca_car_mobility <- PCA(Mobility_car_Norm, quali.sup = 1, graph = FALSE) 
pca_pt_mobility <- PCA(Mobility_pt_Norm, quali.sup = 1, graph = FALSE)
pca_bike_mobility <- PCA(Mobility_bike_Norm, quali.sup = 1, graph = FALSE)
pca_walk_mobility <- PCA(Mobility_walk_Norm, quali.sup = 1, graph = FALSE)

pca_car_safety <- PCA(Safety_car_Norm, quali.sup = 1, graph = FALSE)
pca_bike_safety <- PCA(Safety_bike_Norm, quali.sup = 1, graph = FALSE)
pca_walk_safety <- PCA(Safety_walk_Norm, quali.sup = 1, graph = FALSE)


####### Obtain the variance explained by PCA1 for each dimension and mode

# Define the list of PCA objects you have created
pca_list <- list(
  
  Accessibility_GB = pca_access,
  #Affordability_GB_post_nav = pca_afford_post_nav,
  #Affordability_GB_pre_nav = pca_afford_pre_nav,
  Mobility_GB = pca_mobility,
  Safety_GB = pca_safety,
  Accessibility_car = pca_car_access,
  Accessibility_pt = pca_pt_access,
  Accessibility_bike = pca_bike_access,
  Accessibility_walk = pca_walk_access,
  #Affordability_car = car_afford,
  #Affordability_pt_pre_nav = pt_afford_pre_nav,
  #Affordability_pt_post_nav = pt_afford_post_nav,
  Mobility_car = pca_car_mobility,
  Mobility_pt = pca_pt_mobility,
  Mobility_bike = pca_bike_mobility,
  Mobility_walk = pca_walk_mobility,
  Safety_car = pca_car_safety,
  Safety_bike = pca_bike_safety,
  Safety_walk = pca_walk_safety

  # Add all your other PCA objects here...
)

# Function to extract variance and variable correlations
get_pca_diagnostics <- function(pca_obj, name) {
  
  # 1. Get Variance of PC1
  var_pc1 <- pca_obj$eig[1, 2]
  
  # 2. Get Correlations of variables with PC1
  cors <- as.data.frame(pca_obj$var$cor[, 1])
  colnames(cors) <- "Correlation_PC1"
  cors$Variable <- rownames(cors)
  cors$Dimension_Mode <- name
  cors$PC1_Variance_Percent <- var_pc1
  
  return(cors)
}

# Apply the function to all PCAs and combine into one table
pca_variance_report <- do.call(rbind, lapply(names(pca_list), function(x) {
  get_pca_diagnostics(pca_list[[x]], x)
}))

# Reorder columns for readability
pca_variance_report <- pca_variance_report[, c("Dimension_Mode", "PC1_Variance_Percent", "Variable", "Correlation_PC1")]

######## Write the results to export PCA variances

write_csv(pca_variance_report, "PCA_Variance_and_Correlations.csv")


pca_summary_table <- data.frame(
  Dimension_Mode = names(pca_list),
  PC1_Variance_Percent = sapply(pca_list, function(x) x$eig[1, 2]),
  Eigenvalue_PC1 = sapply(pca_list, function(x) x$eig[1, 1])
)

# 3. View the clean summary
print(pca_summary_table)

#### Export CSV summary

write_csv(pca_summary_table, "PCA_Variance_Table.csv")

# Check the direction of the first variable for each PCA
# 1. Run the loop and store it in an object
direction_results <- do.call(rbind, lapply(names(pca_list), function(x) {
  # Get the correlations for the 1st component
  cors <- pca_list[[x]]$var$cor[, 1]
  
  # Find the variable with the strongest relationship to PC1
  top_idx <- which.max(abs(cors))
  top_var <- names(cors)[top_idx]
  cor_val <- cors[top_idx]
  
  data.frame(
    Dimension = x, 
    Top_Variable = top_var, 
    Correlation = round(cor_val, 3)
  )
}))

# 2. Explicitly print it to the console
print(direction_results)

# 1. Extract Accessibility Score
acc_car_scores <- data.frame(
  dtmnfr = Accessibility_car_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(as.numeric(pca_car_access$ind$coord[, 1]))
)

acc_pt_scores <- data.frame(
  dtmnfr = Accessibility_pt_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(as.numeric(pca_pt_access$ind$coord[, 1]))
)

acc_bike_scores <- data.frame(
  dtmnfr = Accessibility_bike_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(as.numeric(pca_bike_access$ind$coord[, 1]))
)

acc_walk_scores <- data.frame(
  dtmnfr = Accessibility_walk_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(as.numeric(pca_walk_access$ind$coord[, 1]))
)

# 2. Extract Affordability Score
aff_car_scores <- data.frame(
  dtmnfr = Affordability_car_Norm$dtmnfr,
  Affordability_Index = scale_0_100(as.numeric(pca_car_afford$ind$coord[, 1]))
)

# aff_pt_scores <- data.frame(
#   dtmnfr = Affordability_pt_Norm$dtmnfr,
#   Affordability_Index = scale_0_100(as.numeric(pca_pt_afford$ind$coord[, 1]))
# )
# aff_bike_scores <- data.frame(
#   dtmnfr = Affordability_bike_Norm$dtmnfr,
#   Affordability_Index = scale_0_100(as.numeric(pca_bike_afford$ind$coord[, 1]))
# )


# 3. Extract Safety Score
#saf_car_scores <- data.frame(
 #dtmnfr = Safety_car_Norm$dtmnfr,
  #Safety_Index = scale_0_100(pca_car_safety$ind$coord[, 1])
#)
#saf_bike_scores <- data.frame(
 #dtmnfr = Safety_bike_Norm$dtmnfr,
  #Safety_Index = scale_0_100(pca_bike_safety$ind$coord[, 1])
#)
#saf_walk_scores <- data.frame(
 #dtmnfr = Safety_walk_Norm$dtmnfr,
  #Safety_Index = scale_0_100(pca_walk_safety$ind$coord[,1])
#)
#4. Extract Mobility Score
#mob_car_scores <- data.frame(
 # dtmnfr = Mobility_car_Norm$dtmnfr,
  #Mobility_Index = scale_0_100(pca_car_mobility$ind$coord[,1])
#)

######### Safety per mode using avg###########

Safety_car_Norm <- Safety_car_Norm %>%
  mutate(indice_gravidade_carro = 100 - indice_gravidade_carro)

Safety_bike_Norm <- Safety_bike_Norm %>%
  mutate(indice_gravidade_bicicleta = 100 - indice_gravidade_bicicleta)

Safety_walk_Norm <- Safety_walk_Norm %>%
  mutate(indice_gravidade_peoes = 100 - indice_gravidade_peoes)

saf_car_scores <- Safety_car_Norm |> 
  mutate(Safety_Index = indice_gravidade_carro*0.5 + veh_motorizado * 0.5 ) |> 
  select(dtmnfr, Safety_Index)

saf_bike_scores <- Safety_bike_Norm |> 
  mutate(Safety_Index = indice_gravidade_bicicleta*0.5 + veh_bicicleta * 0.5 ) |> 
  select(dtmnfr, Safety_Index)

saf_walk_scores <- Safety_walk_Norm |>
  mutate(Safety_Index = indice_gravidade_peoes*0.5 + veh_peoes * 0.5 ) |>
  select(dtmnfr, Safety_Index)

mob_pt_scores <- data.frame(
  dtmnfr = Mobility_pt_Norm$dtmnfr,
  Mobility_Index = scale_0_100(pca_pt_mobility$ind$coord[,1])
)

mob_bike_scores <- data.frame(
  dtmnfr = Mobility_bike_Norm$dtmnfr,
  Mobility_Index = scale_0_100(pca_bike_mobility$ind$coord[,1])
)
mob_walk_scores <- data.frame(
  dtmnfr = Mobility_walk_Norm$dtmnfr,
  Mobility_Index = scale_0_100(pca_walk_mobility$ind$coord[,1])
)

#### Join Results scores

Final_Results_Master_car <- acc_car_scores |> 
  left_join(car_afford |> rename(Affordability_Index = h_transp_inc_car), by = "dtmnfr") |> 
  left_join(mob_car_scores, by = "dtmnfr") |>
  left_join(saf_car_scores, by = "dtmnfr")

Final_Results_Master_pt <- acc_pt_scores |>
  left_join(pt_afford_post_nav |> rename(Affordability_Index_nav = h_transp_inc_pt), by = "dtmnfr") |> 
  left_join(pt_afford_pre_nav |> rename(Affordability_Index_no_nav = h_transp_inc_pt), by = "dtmnfr") |>
  left_join(mob_pt_scores, by = "dtmnfr") 

Final_Results_Master_bike <- acc_bike_scores |>
  left_join(mob_bike_scores, by = "dtmnfr") |>
  left_join(saf_bike_scores, by = "dtmnfr")

Final_Results_Master_walk <- acc_walk_scores |>
  left_join(mob_walk_scores, by = "dtmnfr") |>
  left_join(saf_walk_scores, by = "dtmnfr")


#Extract the weights from the first principal component 


#pca_weights_access <- pca_access$var$coord[, 1] / sum(pca_access$var$coord[, 1])
#pca_weights_afford <- pca_afford$var$coord[, 1] / sum(pca_afford$var$coord[, 1])
#pca_weights_mobility <- pca_mobility$var$coord[, 1] / sum(pca_mobility$var$coord[, 1])
#pca_weights_safety <- pca_safety$var$coord[, 1] / sum(pca_safety$var$coord[, 1])


# Write the PCA scores to CSV files

write_csv(acc_car_scores, "Accessibility_PCA_car_Scores.csv")
write_csv(acc_pt_scores, "Accessibility_PCA_pt_Scores.csv")
write_csv(acc_bike_scores, "Accessibility_PCA_bike_Scores.csv")
write_csv(acc_walk_scores, "Accessibility_PCA_walk_Scores.csv")

write_csv(aff_car_scores, "Affordability_car_Scores.csv")
write_csv(aff_pt_scores, "Affordability_pt_Scores_pre_nav.csv")
write_csv(aff_pt_scores, "Affordability_pt_Scores_post_nav.csv")


write_csv(saf_car_scores, "Safety_car_Scores.csv")
write_csv(saf_bike_scores, "Safety_bike_Scores.csv")
write_csv(saf_walk_scores, "Safety_walk_Scores.csv")

write_csv(mob_car_scores, "Mobility_PCA_car_Scores.csv")
write_csv(mob_pt_scores, "Mobility_PCA_pt_Scores.csv")
write_csv(mob_bike_scores, "Mobility_PCA_bike_Scores.csv")
write_csv(mob_walk_scores, "Mobility_PCA_walk_Scores.csv")


```



### Step 4: Compute the overall IMPT score

```{r}

# ### EQUAL WEIGHTS
# 
# Results_Equal_Weights <- Results_Equal_Weights |>
#   mutate(
#     # Do the shifting and the math all in one go
#     IMPT_Equal_Weights_geom = ((Accessibility + 0.01) * (Affordability + 0.01) * (Mobility      + 0.01) * (Safety        + 0.01))^(1/4)
#   ) |>
#   mutate(
#     IMPT_Equal_Weights_avg = (Accessibility + Affordability + Mobility + Safety) / 4
#   ) 
# 

                                    
################################################### Compute overall IMPT score using PCA weights + AVG and geometric mean
Final_Results_Master <- Final_Results_Master |> 
  mutate(
    # Adding +1 ensures no zeros, then we subtract it at the end 
    IMPT_score_pca_geom_nav = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (MS_AFF_POST_NAV + 1) * (Safety_Index + 1))^(1/4) - 1
  ) |> 
  mutate(
  IMPT_score_pca_geom_no_nav = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (MS_AFF_PRE_NAV + 1) * (Safety_Index + 1))^(1/4) - 1
  ) |>
mutate( 
  IMPT_score_pca_avg_no_nav = ((Accessibility_Index) + (Mobility_Index) + (MS_AFF_PRE_NAV) + (Safety_Index))/4) |>
  mutate( 
    IMPT_score_pca_avg_nav = ((Accessibility_Index) + (Mobility_Index) + (MS_AFF_POST_NAV) + (Safety_Index))/4)


# Compute overall IMPT with entropy weights
# Assuming 'entropy_weights' is a vector of weights derived from the entropy method for each dimension

compute_entropy_weights <- function(df) {
  # Add 0.0001 to avoid log(0)
  df_shifted <- df + 0.0001
  
  # Calculate proportions
  p_ij <- sweep(df_shifted, 2, colSums(df_shifted), "/")
  
  # Entropy calculation
  n <- nrow(df)
  k <- 1 / log(n)
  
  # Calculate e_j, using 0 for cases where p_ij is 0 to avoid log issues
  e_j <- -k * colSums(p_ij * log(p_ij + 1e-10))
  
  d_j <- 1 - e_j
  w_j <- d_j / sum(d_j)
  
  return(w_j)
}

# 1. dataframe containing only the scores you want to weight
scores_for_entropy1 <- Final_Results_Master |> 
  select(Accessibility_Index, MS_AFF_POST_NAV, Mobility_Index, Safety_Index)

scores_for_entropy2 <- Final_Results_Master |> 
  select(Accessibility_Index, MS_AFF_PRE_NAV, Mobility_Index, Safety_Index)

Entropy_weights_pca_nav <- compute_entropy_weights(scores_for_entropy1)
Entropy_weights_pca_nonav <- compute_entropy_weights(scores_for_entropy2)

# 3. Compute the Index and final results

# Using a +1 shift is very common for 0-100 scales
Final_Results_Master <- Final_Results_Master |>
  mutate(
    IMPT_entropy_pca_nav = (
      (Accessibility_Index + 1)^Entropy_weights_pca_nav[1] *
      (MS_AFF_POST_NAV + 1)^Entropy_weights_pca_nav[2] *
      (Mobility_Index      + 1)^Entropy_weights_pca_nav[3] *
      (Safety_Index        + 1)^Entropy_weights_pca_nav[4]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  ) |> 
  mutate(
    IMPT_entropy_pca_no_nav = (
      (Accessibility_Index + 1)^Entropy_weights_pca_nonav[1] *
      (MS_AFF_PRE_NAV + 1)^Entropy_weights_pca_nonav[2] *
      (Mobility_Index      + 1)^Entropy_weights_pca_nonav[3] *
      (Safety_Index        + 1)^Entropy_weights_pca_nonav[4]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  )



######### Save the final results

write_csv(Results_Equal_Weights, "IMPT_Equal_Weights.csv")
write_csv(Final_Results_Master, "IMPT_PCA_and_Entropy_Scores.csv")

```


#### Compute IMPT per mode

```{r}


                                    
################################################### Compute overall IMPT score using PCA weights + AVG and geometric mean
Final_Results_Master_car <- Final_Results_Master_car |> 
  mutate(
    # Adding +1 ensures no zeros, then we subtract it at the end 
    IMPT_score_pca_geom_car = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Affordability_Index + 1) * (Safety_Index + 1))^(1/4) - 1
  ) |> 
  mutate( 
    IMPT_score_pca_avg_car = ((Accessibility_Index) + (Mobility_Index) + (Affordability_Index) + (Safety_Index))/4)


Final_Results_Master_pt <- Final_Results_Master_pt |> 
  mutate(
    # Adding +1 ensures no zeros, then we subtract it at the end 
    IMPT_score_pca_geom_pt_no_nav = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Affordability_Index_no_nav + 1))^(1/3) - 1
  ) |> 
  mutate (
  IMPT_score_pca_geom_pt_nav = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Affordability_Index_nav + 1))^(1/3) - 1
  ) |>
  mutate( 
    IMPT_score_pca_avg_pt_no_nav = ((Accessibility_Index) + (Mobility_Index) + (Affordability_Index_no_nav))/3) |>
  mutate(
    IMPT_score_pca_avg_pt_nav = ((Accessibility_Index) + (Mobility_Index) + (Affordability_Index_nav))/3)

Final_Results_Master_bike <- Final_Results_Master_bike |>
  mutate(
    # Adding +1 ensures no zeros, then we subtract it at the end 
    IMPT_score_pca_geom_bike = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Safety_Index + 1))^(1/3) - 1
  ) |> 
  mutate( 
    IMPT_score_pca_avg_bike = ((Accessibility_Index) + (Mobility_Index) + (Safety_Index))/4)

Final_Results_Master_walk <- Final_Results_Master_walk |>
  mutate(
    # Adding +1 ensures no zeros, then we subtract it at the end
    IMPT_score_pca_geom_walk = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Safety_Index + 1))^(1/3) - 1
  ) |>
  mutate( 
    IMPT_score_pca_avg_walk = ((Accessibility_Index) + (Mobility_Index) + (Safety_Index))/3)

# Compute overall IMPT with entropy weights


# 1. dataframe containing only the scores you want to weight
scores_for_entropy_car <- Final_Results_Master_car |> 
  select(Accessibility_Index, Affordability_Index, Mobility_Index, Safety_Index)

Entropy_weights_pca_car <- compute_entropy_weights(scores_for_entropy_car)

scores_for_entropy_pt_post_nav <- Final_Results_Master_pt |> 
  select(Accessibility_Index, Affordability_Index_nav, Mobility_Index)

scores_for_entropy_pt_pre_nav <- Final_Results_Master_pt |> 
  select(Accessibility_Index, Affordability_Index_no_nav, Mobility_Index)

Entropy_weights_pca_pt_nav <- compute_entropy_weights(scores_for_entropy_pt_post_nav)

Entropy_weights_pca_pt_no_nav <- compute_entropy_weights(scores_for_entropy_pt_pre_nav)

scores_for_entropy_bike <- Final_Results_Master_bike |> 
  select(Accessibility_Index, Mobility_Index, Safety_Index)
Entropy_weights_pca_bike <- compute_entropy_weights(scores_for_entropy_bike)

scores_for_entropy_walk <- Final_Results_Master_walk |> 
  select(Accessibility_Index, Mobility_Index, Safety_Index)
Entropy_weights_pca_walk <- compute_entropy_weights(scores_for_entropy_walk)

# 3. Compute the Index and final results

# Using a +1 shift is very common for 0-100 scales
Final_Results_Master_car <- Final_Results_Master_car |>
  mutate(
    IMPT_entropy_pca_car = (
      (Accessibility_Index + 1)^Entropy_weights_pca_car[1] *
      (Affordability_Index + 1)^Entropy_weights_pca_car[2] *
      (Mobility_Index      + 1)^Entropy_weights_pca_car[3] *
      (Safety_Index        + 1)^Entropy_weights_pca_car[4]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  ) 

Final_Results_Master_pt <- Final_Results_Master_pt |>
  mutate(
    IMPT_entropy_pca_pt_nav = (
      (Accessibility_Index + 1)^Entropy_weights_pca_pt_nav[1] *
      (Affordability_Index_nav + 1)^Entropy_weights_pca_pt_nav[2] *
      (Mobility_Index      + 1)^Entropy_weights_pca_pt_nav[3]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  ) |> 
  mutate(
    IMPT_entropy_pca_pt_no_nav = (
      (Accessibility_Index + 1)^Entropy_weights_pca_pt_no_nav[1] *
      (Affordability_Index_no_nav + 1)^Entropy_weights_pca_pt_no_nav[2] *
      (Mobility_Index      + 1)^Entropy_weights_pca_pt_no_nav[3]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  )

Final_Results_Master_bike <- Final_Results_Master_bike |>
  mutate(
    IMPT_entropy_pca_bike = (
      (Accessibility_Index + 1)^Entropy_weights_pca_bike[1] *
      (Mobility_Index      + 1)^Entropy_weights_pca_bike[2] *
      (Safety_Index        + 1)^Entropy_weights_pca_bike[3]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  )

Final_Results_Master_walk <- Final_Results_Master_walk |>
  mutate(
    IMPT_entropy_pca_walk = (
      (Accessibility_Index + 1)^Entropy_weights_pca_walk[1] *
      (Mobility_Index      + 1)^Entropy_weights_pca_walk[2] *
      (Safety_Index        + 1)^Entropy_weights_pca_walk[3]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  )


######### Save the final results


write_csv(Final_Results_Master_car, "IMPT_PCA_and_Entropy_Scores_car.csv")
write_csv(Final_Results_Master_pt, "IMPT_PCA_and_Entropy_Scores_pt.csv")
write_csv(Final_Results_Master_bike, "IMPT_PCA_and_Entropy_Scores_bike.csv")
write_csv(Final_Results_Master_walk, "IMPT_PCA_and_Entropy_Scores_walk.csv")

```

# Visualize

```{r}
library(mapview)
freguesias # Read from data_load.R

DATA_LOCATION = "https://impt.server.ushift.pt" # When running locally, get data from remote server
API_KEY = Sys.getenv("IMPT_DATA_KEY") # Set it using usethis::edit_r_environ(), followed by CTRL+F10

IMPT_URL = function(path) {
  # If does not start with "/", add it
  if (!startsWith(path, "/")) {
    path = paste0("/", path)
  }
  local_path = sprintf("%s%s", DATA_LOCATION, path)
  # If data location starts with "http", add api key to url
  if (startsWith(DATA_LOCATION, "http")) {
    # If API_KEY empty or not defined, throw error
    if (API_KEY == "") {
      stop("IMPT_DATA_KEY env var is not defined. Please set it using usethis::edit_r_environ() and restart R.")
    }
    return(sprintf("%s%s?key=%s", DATA_LOCATION, path, API_KEY))
  } else {
    # If local folder does not exist, create, recursively if needed
    if (!dir.exists(dirname(local_path))) {
      dir.create(dirname(local_path), recursive = TRUE)
    }
  }
  # Otherwise, return local path
  return(local_path)
}


# IMPT PCA and Entropy scores
impt_pca = read.csv(IMPT_URL("results_aggregated/20260330/IMPT_PCA_and_Entropy_Scores.csv"))
impt_pca_sf = freguesias |>
  select(-nuts2, -area_ha) |>
  mutate(dtmnfr = as.integer(dtmnfr)) |>
  left_join(
    impt_pca |>
      rename(
        Accessibility_PCA = Accessibility_Index,
        Affordability_PCA_prenav = MS_AFF_PRE_NAV,
        Affordability_PCA_postnav = MS_AFF_POST_NAV,
        Mobility_PCA = Mobility_Index,
        Safety_PCA = Safety_Index
      )
  ) |> 
  # Round PCA scores to 2 decimals
  mutate(
    across(starts_with("Accessibility"), ~ round(., 2)),
    across(starts_with("Affordability"), ~ round(., 2)),
    across(starts_with("Mobility"), ~ round(., 2)),
    across(starts_with("Safety"), ~ round(., 2)),
    across(starts_with("IMPT"), ~ round(., 2))
  ) |>
  # Columns order
  select(
    dtmnfr, freguesia, municipio, 
    starts_with("Accessibility"), starts_with("Affordability"), starts_with("Mobility"), starts_with("Safety"),
    starts_with("IMPT")
  )
# View(impt_pca_sf |> st_drop_geometry())

mapview(impt_pca_sf, zcol="Accessibility_PCA", layer.name="PCA Accessibility", hide=TRUE) +
  mapview(impt_pca_sf, zcol="Affordability_PCA_prenav", layer.name="PCA Affordability - single fare", hide=TRUE) +
  mapview(impt_pca_sf, zcol="Affordability_PCA_postnav", layer.name="PCA Affordability - navegante", hide=TRUE) +
  mapview(impt_pca_sf, zcol="Mobility_PCA", layer.name="PCA Mobility", hide=TRUE) +
  mapview(impt_pca_sf, zcol="Safety_PCA", layer.name="PCA Safety", hide=TRUE) +
  
  mapview(impt_pca_sf, zcol="IMPT_score_pca_geom_nav", layer.name="IMPT PCA Geom - navegante", hide=TRUE) + 
  mapview(impt_pca_sf, zcol="IMPT_score_pca_geom_no_nav", layer.name="IMPT PCA Geom - single fare", hide=TRUE) +
  mapview(impt_pca_sf, zcol="IMPT_score_pca_avg_nav", layer.name="IMPT PCA Avg - navegante", hide=TRUE) + 
   mapview(impt_pca_sf, zcol="IMPT_score_pca_avg_no_nav", layer.name="IMPT PCA Avg - single fare", hide=TRUE) +
  mapview(impt_pca_sf, zcol="IMPT_entropy_pca_nav", layer.name="IMPT PCA Entropy - navegante", hide=TRUE) +
  mapview(impt_pca_sf, zcol="IMPT_entropy_pca_no_nav", layer.name="IMPT PCA Entropy - single fare", hide=TRUE)
```

