---
title: "Accessibility (parish)"
output: html_document
---

> This script assumes variables at `Global.Rmd` have been loaded

```{r}
library(readxl)
library(dplyr)
library(mapview)
```

# Indicators

## Number of jobs reachable in 15 minutes

```{r}
data_jobs_reachable = read_excel("/data/IMPT/BaseDados_PMMUS/3-CondiçõesAcesso/3\ -\ Condições\ Acesso.xlsx", sheet="trabalho")
View(data_jobs_reachable)

parish_jobs_reachable  = data_jobs_reachable |>
  left_join(grid_hex_pop |> select(ID, N_INDIVIDU) |> sf::st_drop_geometry(), by=c("id"="ID")) |>
  group_by(concelho, freg) |> 
  summarise(jobs_reachable = weighted.mean(WALK, N_INDIVIDU))

# Normalize using min-max normalization
parish_access_jobs = parish_jobs_reachable |>
  mutate(access = (jobs_reachable - min(jobs_reachable)) / (max(jobs_reachable) - min(jobs_reachable)))
View(parish_access_jobs)
```

## Travel time to nearest school

```{r}
data_nearest_school = read_excel("/data/IMPT/BaseDados_PMMUS/3-CondiçõesAcesso/3\ -\ Condições\ Acesso.xlsx", sheet="p_esc") |>
  mutate(WALK = as.numeric(WALK))
View(data_nearest_school)

parish_time_school = data_nearest_school |>
  left_join(grid_hex_pop |> select(ID, N_INDIVIDU) |> sf::st_drop_geometry(), by=c("id"="ID")) |>
  group_by(concelho, freg) |> 
  summarise(time_school = weighted.mean(WALK, N_INDIVIDU, na.rm=TRUE))

parish_ratio_access_school = data_nearest_school |>
  left_join(grid_hex_pop |> select(ID, N_INDIVIDU) |> sf::st_drop_geometry(), by=c("id"="ID")) |>
  mutate(accessible = ifelse(is.na(WALK), 0, 1)) |>
  group_by(concelho, freg) |> 
  summarise(ratio_access_school = sum(accessible * N_INDIVIDU) / sum(N_INDIVIDU))

# Normalize combining time and ratio of access
parish_access_school = parish_time_school |>
  left_join(parish_ratio_access_school, by=c("concelho", "freg")) |>
  mutate(access = ratio_access_school * (1-time_school/15)) |>
  mutate(access = ifelse(is.na(access), 0, access))
```


## Travel time to primary health care

```{r}
data_nearest_health = read_excel("/data/IMPT/BaseDados_PMMUS/3-CondiçõesAcesso/3\ -\ Condições\ Acesso.xlsx", sheet="csp") |>
  mutate(WALK = as.numeric(WALK))
View(data_nearest_health)

parish_time_health = data_nearest_health |>
  left_join(grid_hex_pop |> select(ID, N_INDIVIDU) |> sf::st_drop_geometry(), by=c("id"="ID")) |>
  group_by(concelho, freg) |> 
  summarise(time_health = weighted.mean(WALK, N_INDIVIDU, na.rm=TRUE))

parish_ratio_access_health = data_nearest_health |>
  left_join(grid_hex_pop |> select(ID, N_INDIVIDU) |> sf::st_drop_geometry(), by=c("id"="ID")) |>
  mutate(accessible = ifelse(is.na(WALK), 0, 1)) |>
  group_by(concelho, freg) |> 
  summarise(ratio_access_health = sum(accessible * N_INDIVIDU) / sum(N_INDIVIDU))

parish_access_health = parish_time_health |>
  left_join(parish_ratio_access_health, by=c("concelho", "freg")) |>
  # (1-time_health/15) converts time to a 0-1 scale (0 is 15 minutes or more, 1 is 0 minutes)
  mutate(access = ratio_access_health * (1-time_health/15)) |> 
  mutate(access = ifelse(is.na(access), 0, access))
```


## Travel time to supermarket

```{r}
data_nearest_supermarket = read_excel("/data/IMPT/BaseDados_PMMUS/3-CondiçõesAcesso/3\ -\ Condições\ Acesso.xlsx", sheet="superm") |>
  mutate(WALK = as.numeric(WALK))
View(data_nearest_supermarket)

parish_time_supermarket = data_nearest_supermarket |>
  left_join(grid_hex_pop |> select(ID, N_INDIVIDU) |> sf::st_drop_geometry(), by=c("id"="ID")) |>
  group_by(concelho, freg) |> 
  summarise(time_supermarket = weighted.mean(WALK, N_INDIVIDU, na.rm=TRUE))

parish_ratio_access_supermarket = data_nearest_supermarket |>
  left_join(grid_hex_pop |> select(ID, N_INDIVIDU) |> sf::st_drop_geometry(), by=c("id"="ID")) |>
  mutate(accessible = ifelse(is.na(WALK), 0, 1)) |>
  group_by(concelho, freg) |> 
  summarise(ratio_access_supermarket = sum(accessible * N_INDIVIDU) / sum(N_INDIVIDU))

parish_access_supermarket = parish_time_supermarket |>
  left_join(parish_ratio_access_supermarket, by=c("concelho", "freg")) |>
  mutate(access = ratio_access_supermarket * (1-time_supermarket/15)) |>
  mutate(access = ifelse(is.na(access), 0, access))
```

# Factor weighting

```{r}
parish_accessibility = parish_access_jobs |>
  select(concelho, freg, access_jobs = access) |>
  left_join(parish_access_school |> select(concelho, freg, access_school = access), by=c("concelho", "freg")) |>
  left_join(parish_access_health |> select(concelho, freg, access_health = access), by=c("concelho", "freg")) |>
  left_join(parish_access_supermarket |> select(concelho, freg, access_supermarket = access), by=c("concelho", "freg"))
```


## Equal weights

Assign equal weights to all accessibility indicators

## PCA

Perform PCA to derive weights from data

> Based on https://www.datacamp.com/tutorial/pca-analysis-r

```{r}
# PCA for accessibility indicators
library(factoextra)

numerical_data = parish_accessibility[,3:6]
head(numerical_data)

data.pca <- prcomp(numerical_data, center=TRUE, scale.=FALSE) # scale not required as data is already normalized
summary(data.pca)

-data.pca$rotation
pca_scores = -data.pca$x[,1]
pca_scores = (pca_scores - min(pca_scores)) / (max(pca_scores) - min(pca_scores))
```

```{r}
data.pca$loadings[, 1:2]
# Scree Plot: Percentage of variance explained by each component
fviz_eig(data.pca, addlabels = TRUE)

# Biplot: Similarities and dissimilarities between the samples (groups) ab impact of each attribute on each of the principal components
fviz_pca_var(data.pca, col.var = "black")
fviz_pca_biplot(data.pca, label = "var", addEllipses=TRUE, ellipse.level=0.95,ggtheme = theme_minimal())

# Cos2: How much each variable is represented in a given component
fviz_cos2(data.pca, choice = "var", axes = 1:2)

# Biplot + Cos2
fviz_pca_var(data.pca, col.var = "cos2",
            gradient.cols = c("black", "orange", "green"),
            repel = TRUE)
```



## Final accessibility score

```{r}
accessibility_score = parish_accessibility |>
  mutate(
    # Equal weights
    ACCESS_equal_weight = (access_jobs + access_school + access_health + access_supermarket) / 4
  )

accessibility_score$ACCESS_pca = pca_scores

View(accessibility_score)
```

```{r}
library(stringi)

accessibility_score_sf = accessibility_score |>
  mutate(
    concelho_match = tolower(stri_trans_general(concelho, "Latin-ASCII")),
    freg_match = tolower(stri_trans_general(freg, "Latin-ASCII"))
  ) |>
  left_join(parishes |> mutate(
    concelho_match = tolower(stri_trans_general(Concelho, "Latin-ASCII")),
    freg_match = tolower(stri_trans_general(Freguesia, "Latin-ASCII"))
  ), by=c("concelho_match", "freg_match")) |>
  select(-concelho_match, -freg_match, -Concelho, -Freguesia) |>
  sf::st_as_sf()

map = mapview::mapview(accessibility_score_sf, zcol="ACCESS_equal_weight", layer.name="Accessibility Score<br/>(equal weights)") +
  mapview::mapview(accessibility_score_sf, zcol="ACCESS_pca", layer.name=sprintf(
    "Accessibility Score<br/>(PCA, %.2f%% var explained)",
    summary(data.pca)$importance[2,1]*100
    ))
map

# mapshot(map, "outputs/accessibility.html", selfcontained = FALSE)
```


