---
title: "IMPT"
output: html_document
date: "2026-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# Load all necessary libraries here
library(dplyr)
library(readr)
library(ggplot2)
library(sf) # If you are using spatial data for Freguesias
library(tidyr)
library(FactoMineR)
library(factoextra)
library(stringr) # For string manipulation, if neededring)
library(janitor) # For data cleaning, if needed
```



## The code to compute the IMPT


### Leaving out Mobility costs for now

######### Data prepatation #########################

```{r}
# Data loading Accessibility
freguesia_accessibility <- read_csv("/data/IMPT/results_data/freguesia_accessibility.csv")
Accessibility <- freguesia_accessibility |> select(-nuts)
```

```{r}
# Data loading Mobility
freguesias_infrastructure_ratio <- read_csv("/data/IMPT/mobility/freguesias_infrastructure_ratio.csv")
# names(freguesias_infrastructure_ratio)
freguesias_infrastructure_ratio <- freguesias_infrastructure_ratio |>
  select(
    dtmnfr, road_length, pedpath_length, cycleway_length,
    segregated_cycleway_length, pedpath_to_road_ratio, cycleway_to_road_ratio,
    cycling_quality_ratio
  )

freguesias_stops_coverage <- read_csv("/data/IMPT/mobility/freguesias_stops_coverage.csv")
# names(freguesias_stops_coverage)
#### There is not a varaible named "Stops_coverage", and the on at mauricio drive has different values than  the suspected one
# summary(Mobility2$Stops_coverage)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#  0.1337  0.7302  0.9211  0.8522  0.9833  1.0000
# summary(freguesias_stops_coverage$ratio_served_population)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#  0.3500  0.9000  0.9800  0.9354  1.0000  1.0000

freguesias_stops_coverage <- freguesias_stops_coverage |>
  select(freguesia, ratio_served_population) # will use this for now...
# we need to add the dtmnfr code to be able to join
freguesias_names <- st_read("/data/IMPT/geo/freguesias_2024_unique.gpkg") |>
  st_drop_geometry() |>
  select(dtmnfr, freguesia)
freguesias_stops_coverage <- freguesias_stops_coverage |>
  left_join(freguesias_names, by = "freguesia") |>
  select(-freguesia) |>
  mutate(dtmnfr = as.numeric(dtmnfr)) |>
  rename(Stops_coverage = ratio_served_population)

freguesias_shared_mobility <- read_csv("/data/IMPT/mobility/freguesias_shared_mobility.csv")
# names(freguesias_shared_mobility)
freguesias_shared_mobility <- freguesias_shared_mobility |>
  select(dtmnfr, shared_mobility_points) |>
  rename(Shared_mobility = shared_mobility_points)

freguesias_mobility <- read_csv("/data/IMPT/results_data/freguesia_mobility.csv")
names(freguesias_mobility)
freguesias_mobility <- freguesias_mobility |>
  select(-nuts)

freguesias_headways <- read_csv("/data/IMPT/mobility_transit/freguesias_headways.csv")
names(freguesias_headways)
# summary(Mobility2$PT_Waitingtime)
# summary(freguesias_headways$weighted_waiting_time_peak)
# summary(Mobility2$PT_Weight_frequendy_peak_day)
# summary(freguesias_headways$weighted_frequency_day)
# summary(freguesias_headways$weighted_frequency_peak) # we want this one!
freguesias_headways <- freguesias_headways |>
  select(dtmnfr, weighted_frequency_peak, weighted_waiting_time_peak) |>
  rename(
    PT_Waitingtime = weighted_waiting_time_peak,
    PT_Weight_frequency_peak = weighted_frequency_peak
  )

## Join all tables
Mobility <- freguesias_infrastructure_ratio |>
  left_join(freguesias_stops_coverage, by = "dtmnfr") |>
  left_join(freguesias_shared_mobility, by = "dtmnfr") |>
  left_join(freguesias_mobility, by = "dtmnfr") |>
  left_join(freguesias_headways, by = "dtmnfr")

# names(Mobility)
# names(Mobility2)
```

```{r}
# Data loading Safety
accidents_by_freguesia_5years_dentrolocalidades <- read_csv("/data/IMPT/safety/accidents_by_freguesia_5years_dentrolocalidades.csv")
# names(accidents_by_freguesia_5years_dentrolocalidades)
Safety <- accidents_by_freguesia_5years_dentrolocalidades |>
  select(-freguesia, -mun_id, -population) |>
  rename(dtmnfr = freg_id) |>
  mutate(across(where(is.numeric), ~ ifelse(is.na(.x), 0, .x))) # Replace NA with 0 for safety dataset
```

```{r}
# Data loading Affordability
freguesia_affordability_mob <- read_csv("/data/IMPT/results_data/freguesia_affordability.csv")
names(freguesia_affordability_mob)
freguesia_income <- read_csv("/data/IMPT/landuse/freguesias_income_housing_gini.csv") |>
  select(-gini_coef) |>
  rename(dtmnfr = freg_id) |>
  mutate(housing_costs_year = housing_costs * 12)
# join both
freguesia_affordability <- freguesia_affordability_mob |>
  left_join(freguesia_income, by = "dtmnfr")

## With Navegante monthly pass
Affordability_navegante <- freguesia_affordability |>
  mutate(h_transp_inc_car = (housing_costs_year + affordability_car_total_money * 250) / income_hh) |> # consider 250 working days
  mutate(h_transp_inc_pt = (housing_costs_year + affordability_transit_pass_total_money * 250) / income_hh) |>
  mutate(transp_inc_car = (affordability_car_total_money * 250) / income_hh) |>
  mutate(transp_inc_pt = (affordability_transit_pass_total_money * 250) / income_hh) |>
  select(dtmnfr, h_transp_inc_car, h_transp_inc_pt, transp_inc_car, transp_inc_pt)
#   mutate(across(where(is.numeric), ~ ifelse(is.na(.x), 0, .x))) # Change Canha

mean_viz_canha <- Affordability_navegante |> # Filter dtmnfr starting with 1507
  filter(str_starts(dtmnfr, "1507")) |>
  filter(dtmnfr != "150701") |>
  summarise(
    mean_h_transp_inc_pt = mean(h_transp_inc_pt, na.rm = TRUE),
    mean_transp_inc_pt = mean(transp_inc_pt, na.rm = TRUE)
  )

#### Replace zeros with the municipality avg
Affordability_navegante <- Affordability_navegante |>
  mutate(h_transp_inc_pt = case_when(
    dtmnfr == "150701" ~ mean_viz_canha$mean_h_transp_inc_pt,
    TRUE ~ h_transp_inc_pt
  )) |>
  mutate(transp_inc_pt = case_when(
    dtmnfr == "150701" ~ mean_viz_canha$mean_transp_inc_pt,
    TRUE ~ transp_inc_pt
  ))

## With single fare tickets
Affordability_singlefare <- freguesia_affordability |>
  mutate(h_transp_inc_car = (housing_costs_year + affordability_car_total_money * 250) / income_hh) |>
  mutate(h_transp_inc_pt = (housing_costs_year + affordability_transit_single_fare_total_money * 250) / income_hh) |>
  mutate(transp_inc_car = (affordability_car_total_money * 250) / income_hh) |>
  mutate(transp_inc_pt = (affordability_transit_single_fare_total_money * 250) / income_hh) |>
  select(dtmnfr, h_transp_inc_car, h_transp_inc_pt, transp_inc_car, transp_inc_pt)

mean_viz_canha2 <- Affordability_singlefare |> # Filter dtmnfr starting with 1507
  filter(str_starts(dtmnfr, "1507")) |>
  filter(dtmnfr != "150701") |>
  summarise(
    mean_h_transp_inc_pt = mean(h_transp_inc_pt, na.rm = TRUE),
    mean_transp_inc_pt = mean(transp_inc_pt, na.rm = TRUE)
  )

#### Replace zeros with the municipality avg
Affordability_singlefare <- Affordability_singlefare |>
  mutate(h_transp_inc_pt = case_when(
    dtmnfr == "150701" ~ mean_viz_canha2$mean_h_transp_inc_pt,
    TRUE ~ h_transp_inc_pt
  )) |>
  mutate(transp_inc_pt = case_when(
    dtmnfr == "150701" ~ mean_viz_canha2$mean_transp_inc_pt,
    TRUE ~ transp_inc_pt
  ))


census_modal_share <- read_csv("/data/IMPT/census2021/census_modal_share_parish.csv") |>
  select(dtmnfr, total, pt, private_vehicle, active, pt_share, private_vehicle_share, active_share)
```


######## Chunk for specific mode dimensions datasets


```{r}
###################### Create a dataset per mode per dimension #####################

Accessibility_car <- Accessibility |>
  select(dtmnfr, contains("_car_"))
Accessibility_pt <- Accessibility |>
  select(dtmnfr, contains("_transit_"))
Accessibility_bike <- Accessibility |>
  select(dtmnfr, contains("_bike_"))
Accessibility_walk <- Accessibility |>
  select(dtmnfr, contains("_walk_"))


Mobility_car <- Mobility |>
  select(dtmnfr, contains("_car"), "road_length")
Mobility_pt <- Mobility |>
  select(dtmnfr, contains("PT"), "Stops_coverage", contains("transit"), Shared_mobility)
Mobility_bike <- Mobility |>
  select(dtmnfr, contains("cycl"), contains("bike"))
Mobility_walk <- Mobility |>
  select(dtmnfr, contains("_walk"), contains("ped"))

Affordability_car <- Affordability_navegante |>
  select(dtmnfr, contains("car"))
Affordability_pt_navegante <- Affordability_navegante |>
  select(dtmnfr, contains("pt"))
Affordability_pt_singlefare <- Affordability_singlefare |>
  select(dtmnfr, contains("pt"))


Safety_car <- Safety |>
  select(dtmnfr, contains("car"), contains("motorizado"))
Safety_bike <- Safety |>
  select(dtmnfr, contains("bici"))

# Replace NA with 0 for bike safety

Safety_bike <- Safety_bike |>
  mutate(across(where(is.numeric), ~ ifelse(is.na(.x), 0, .x)))
Safety_walk <- Safety |>
  select(dtmnfr, contains("peo"))
```




### Step 1: Normalize the indicators

Use Normalize benefit for indicators where higher is better and Normalize cost for indicators where lower is better.
```{r cars}
# Use for "More is Better" (e.g., Number of stops, Frequency, Safety score)
normalize_benefit <- function(x) {
  x_min <- min(x, na.rm = TRUE)
  x_max <- max(x, na.rm = TRUE)

  # Standard Min-Max scaled to 1-100
  return(1 + ((x - x_min) / (x_max - x_min)) * (100 - 1))
}

# Use for "More is Worse" (e.g., Travel time, Cost, Distance to station)
normalize_cost <- function(x) {
  x_min <- min(x, na.rm = TRUE)
  x_max <- max(x, na.rm = TRUE)

  # Inverted Min-Max scaled to 1-100
  # This makes the lowest raw value become 100
  return(1 + ((x_max - x) / (x_max - x_min)) * (100 - 1))
}


Accessibility_Norm <- Accessibility |>
  mutate(
    across(starts_with("access"), ~ normalize_benefit(.x)), # Use c() inside starts_with for multiple prefixes
    across(starts_with("mobility"), ~ normalize_cost(.x)) # Same for your negative indicators
  ) |>
  mutate(
    across(where(is.numeric), ~ ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x))
  ) # replace NA with the mean

Mobility_Norm <- Mobility |>
  mutate(
    across(starts_with(c("PT_Wait", "mobility")), ~ normalize_cost(.x)), # Use c() inside starts_with for multiple prefixes
    across(!c(dtmnfr, starts_with(c("PT_Wait", "mobility"))), ~ normalize_benefit(.x)) # Same for the negative indicators
  )

Affordability_Norm_navegante <- Affordability_navegante |> ### All are negative
  mutate(across(!dtmnfr, ~ normalize_cost(.x)))
Affordability_Norm_singlefare <- Affordability_singlefare |> ### All are negative
  mutate(across(!dtmnfr, ~ normalize_cost(.x)))

Safety_Norm <- Safety |> ### All are negative
  mutate(across(!dtmnfr, ~ normalize_cost(.x)))

##### Export normalized dataset

write_csv(Mobility_Norm, "/data/IMPT/impt/norm/Mobility_Norm.csv")
write_csv(Accessibility_Norm, "/data/IMPT/impt/norm/Accessibility_Norm.csv")
write_csv(Affordability_Norm_navegante, "/data/IMPT/impt/norm/Affordability_Norm_navegante.csv")
write_csv(Affordability_Norm_singlefare, "/data/IMPT/impt/norm/Affordability_Norm_singlefare.csv")
write_csv(Safety_Norm, "/data/IMPT/impt/norm/Safety_Norm.csv")
```

### Normalize indicators for per mode datasets

The functions to normalize are in the previous chunk

```{r}
Accessibility_car_Norm <- Accessibility_car |>
  mutate(
    across(starts_with("access"), ~ normalize_benefit(.x)), # Use c() inside starts_with for multiple prefixes
    across(starts_with("mobility"), ~ normalize_cost(.x))
  ) # Same for your negative indicators

Accessibility_pt_Norm <- Accessibility_pt |>
  mutate(
    across(starts_with("access"), ~ normalize_benefit(.x)),
    across(starts_with("mobility"), ~ normalize_cost(.x))
  ) |>
  mutate(
    across(where(is.numeric), ~ ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x))
  ) # replace NA with the mean)

Accessibility_bike_Norm <- Accessibility_bike |>
  mutate(
    across(starts_with("access"), ~ normalize_benefit(.x)),
    across(starts_with("mobility"), ~ normalize_cost(.x))
  )

Accessibility_walk_Norm <- Accessibility_walk |>
  mutate(
    across(starts_with("access"), ~ normalize_benefit(.x)),
    across(starts_with("mobility"), ~ normalize_cost(.x))
  )


Mobility_car_Norm <- Mobility_car |>
  mutate(
    across(starts_with(c("mobility")), ~ normalize_cost(.x)),
    across(!c(dtmnfr, starts_with(c("mobility"))), ~ normalize_benefit(.x))
  )
Mobility_pt_Norm <- Mobility_pt |>
  mutate(
    across(starts_with(c("PT_Wait", "mobility")), ~ normalize_cost(.x)),
    across(!c(dtmnfr, starts_with(c("PT_Wait", "mobility"))), ~ normalize_benefit(.x))
  )
Mobility_bike_Norm <- Mobility_bike |>
  mutate(
    across(starts_with(c("mobility")), ~ normalize_cost(.x)),
    across(!c(dtmnfr, starts_with(c("mobility"))), ~ normalize_benefit(.x))
  )
Mobility_walk_Norm <- Mobility_walk |>
  mutate(
    across(starts_with(c("mobility")), ~ normalize_cost(.x)),
    across(!c(dtmnfr, starts_with(c("mobility"))), ~ normalize_benefit(.x))
  )


Affordability_car_Norm <- Affordability_car |> ### All are negative
  mutate(across(!dtmnfr, ~ normalize_cost(.x)))
Affordability_pt_Norm_navegante <- Affordability_pt_navegante |> ### All are negative
  mutate(across(!dtmnfr, ~ normalize_cost(.x)))
Affordability_pt_Norm_singlefare <- Affordability_pt_singlefare |> ### All are negative
  mutate(across(!dtmnfr, ~ normalize_cost(.x)))

afford_navegante <- Affordability_Norm_navegante |>
  left_join(census_modal_share, by = "dtmnfr") |>
  mutate(MS_CAR_AFF = h_transp_inc_car * private_vehicle_share) |>
  mutate(MS_PT_AFF = h_transp_inc_pt * pt_share) |>
  mutate(MS_AFF_POST_NAV = (MS_CAR_AFF + MS_PT_AFF) / (private_vehicle_share + pt_share + active_share))
afford_singlefare <- Affordability_Norm_singlefare |>
  left_join(census_modal_share, by = "dtmnfr") |>
  mutate(MS_CAR_AFF = h_transp_inc_car * private_vehicle_share) |>
  mutate(MS_PT_AFF = h_transp_inc_pt * pt_share) |>
  mutate(MS_AFF_PRE_NAV = (MS_CAR_AFF + MS_PT_AFF) / (private_vehicle_share + pt_share + active_share))


Safety_car_Norm <- Safety_car |> ### All are negative
  mutate(across(!dtmnfr, ~ normalize_cost(.x)))
Safety_bike_Norm <- Safety_bike |> ### All are negative
  mutate(across(!dtmnfr, ~ normalize_cost(.x)))
Safety_walk_Norm <- Safety_walk |> ### All are negative
  mutate(across(!dtmnfr, ~ normalize_cost(.x)))


##### Export normalized datasets per mode

write_csv(Mobility_car_Norm, "/data/IMPT/impt/norm/Mobility_car_Norm.csv")
write_csv(Mobility_pt_Norm, "/data/IMPT/impt/norm/Mobility_pt_Norm.csv")
write_csv(Mobility_bike_Norm, "/data/IMPT/impt/norm/Mobility_bike_Norm.csv")
write_csv(Mobility_walk_Norm, "/data/IMPT/impt/norm/Mobility_walk_Norm.csv")

write_csv(Accessibility_car_Norm, "/data/IMPT/impt/norm/Accessibility_car_Norm.csv")
write_csv(Accessibility_pt_Norm, "/data/IMPT/impt/norm/Accessibility_pt_Norm.csv")
write_csv(Accessibility_bike_Norm, "/data/IMPT/impt/norm/Accessibility_bike_Norm.csv")
write_csv(Accessibility_walk_Norm, "/data/IMPT/impt/norm/Accessibility_walk_Norm.csv")

write_csv(Affordability_car_Norm, "/data/IMPT/impt/norm/Affordability_car_Norm.csv")
write_csv(Affordability_pt_Norm_singlefare, "/data/IMPT/impt/norm/Affordability_pt_Norm_singlefare.csv")
write_csv(Affordability_pt_Norm_navegante, "/data/IMPT/impt/norm/Affordability_pt_Norm_navegante.csv")

write_csv(Safety_car_Norm, "/data/IMPT/impt/norm/Safety_car_Norm.csv")
write_csv(Safety_bike_Norm, "/data/IMPT/impt/norm/Safety_bike_Norm.csv")
write_csv(Safety_walk_Norm, "/data/IMPT/impt/norm/Safety_walk_Norm.csv")
```


## Step 2: Compute dimensions weights based on indicators

We will use two approaches:

- Equal weights: All indicators contribute equally to the dimension score.

```{r pressure, echo=FALSE}
# Load necessary libraries


# Equal weights
# Function to calculate the Dimension Score (Equal Weights for Indicators)
# calculate_dimension_EW <- function(df_norm) {
#   # We exclude the ID column (dtmnfr) and take the average of all numeric columns
#   score <- rowMeans(df_norm[, -1], na.rm = TRUE)
#
# }
#
# # Calculate the 4 dimensions
# Acc_EW_Score  <- calculate_dimension_EW(Accessibility_Norm)
# Aff_EW_Score  <- calculate_dimension_EW(Affordability_Norm)
# Mob_EW_Score  <- calculate_dimension_EW(Mobility_Norm)
# Saf_EW_Score  <- calculate_dimension_EW(Safety_Norm)
#
# # Combine them into a new Results table
# Results_Equal_Weights <- data.frame(
#   dtmnfr = Accessibility_Norm$dtmnfr,
#   Accessibility = Acc_EW_Score,
#   Affordability = Aff_EW_Score,
#   Mobility      = Mob_EW_Score,
#   Safety        = Saf_EW_Score
# )
```


#### Equal weights per mode

```{r,eval=FALSE}
# Calculate the 4 dimensions
Acc_car_EW_Score <- calculate_dimension_EW(Accessibility_car_Norm)
Acc_pt_EW_Score <- calculate_dimension_EW(Accessibility_pt_Norm)
Acc_bike_EW_Score <- calculate_dimension_EW(Accessibility_bike_Norm)
Acc_walk_EW_Score <- calculate_dimension_EW(Accessibility_walk_Norm)


Aff_car_EW_Score <- calculate_dimension_EW(Affordability_car_Norm)
Aff_pt_EW_Score <- calculate_dimension_EW(Affordability_pt_Norm)
Aff_bike_EW_Score <- calculate_dimension_EW(Affordability_bike_Norm)


Mob_car_EW_Score <- calculate_dimension_EW(Mobility_car_Norm)
Mob_pt_EW_Score <- calculate_dimension_EW(Mobility_pt_Norm)
Mob_bike_EW_Score <- calculate_dimension_EW(Mobility_bike_Norm)
Mob_walk_EW_Score <- calculate_dimension_EW(Mobility_walk_Norm)

Saf_car_EW_Score <- calculate_dimension_EW(Safety_car_Norm)
Saf_bike_EW_Score <- calculate_dimension_EW(Safety_bike_Norm)
Saf_walk_EW_Score <- calculate_dimension_EW(Safety_walk_Norm)

# Combine them into a new Results table
Results_Equal_Weights_car <- data.frame(
  dtmnfr = Accessibility_car_Norm$dtmnfr,
  Accessibility = Acc_car_EW_Score,
  Affordability = Aff_car_EW_Score,
  Mobility = Mob_car_EW_Score,
  Safety = Saf_car_EW_Score
)

Results_Equal_Weights_pt <- data.frame(
  dtmnfr = Accessibility_pt_Norm$dtmnfr,
  Accessibility = Acc_pt_EW_Score,
  Affordability = Aff_pt_EW_Score,
  Mobility = Mob_pt_EW_Score,
  Safety = NA # No safety indicators for PT
)

Results_Equal_Weights_bike <- data.frame(
  dtmnfr = Accessibility_bike_Norm$dtmnfr,
  Accessibility = Acc_bike_EW_Score,
  Affordability = Aff_bike_EW_Score,
  Mobility = Mob_bike_EW_Score,
  Safety = Saf_bike_EW_Score
)

Results_Equal_Weights_walk <- data.frame(
  dtmnfr = Accessibility_walk_Norm$dtmnfr,
  Accessibility = Acc_walk_EW_Score,
  Affordability = NA, # No affordability indicators for walking
  Mobility = Mob_walk_EW_Score,
  Safety = Saf_walk_EW_Score
)
```


### Step 3: Compute USING PCA: We will perform Principal Component Analysis (PCA) to determine the weights based on the variance explained by each indicator. Using PC1.

```{r}
##################### PCA #####################################

# Perform PCA on the normalized indicators
# Assuming 'data' is dataset with normalized indicators

pca_access <- PCA(Accessibility_Norm, quali.sup = 1, graph = FALSE)
# pca_afford_post_nav <- PCA(Affordability_Norm_post_nav, quali.sup = 1, graph = FALSE)
# pca_afford_pre_nav <- PCA(Affordability_Norm_pre_nav, quali.sup = 1, graph = FALSE)
pca_mobility <- PCA(Mobility_Norm, quali.sup = 1, graph = FALSE)
pca_safety <- PCA(Safety_Norm, quali.sup = 1, graph = FALSE)


# A helper function to scale any vector 0-100
scale_0_100 <- function(x) {
  x_min <- min(x, na.rm = TRUE)
  x_max <- max(x, na.rm = TRUE)
  # Logic: 1 + (normalized_value * 99)
  # This ensures the lowest value is 1 and the highest is 100
  1 + ((x - x_min) / (x_max - x_min) * 99)
}

# 1. Extract Accessibility Score
acc_scores <- data.frame(
  dtmnfr = Accessibility_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(pca_access$ind$coord[, 1])
)

# 2. Extract Affordability Score
# aff_scores_post_nav <- data.frame(
#   dtmnfr = Affordability_Norm_post_nav$dtmnfr,
#   Affordability_Index_post_nav = scale_0_100(pca_afford_post_nav$ind$coord[, 1]))
# aff_scores_pre_nav <- data.frame(
#   dtmnfr = Affordability_Norm_pre_nav$dtmnfr,
#   Affordability_Index_pre_nav = scale_0_100(pca_afford_pre_nav$ind$coord[, 1]))
afford_navegante_fin <- afford_navegante |>
  select(dtmnfr, MS_AFF_POST_NAV) |>
  rename(Affordability_Index_navegante = MS_AFF_POST_NAV)
afford_singlefare_fin <- afford_singlefare |>
  select(dtmnfr, MS_AFF_PRE_NAV) |>
  rename(Affordability_Index_singlefare = MS_AFF_PRE_NAV)


# 3. Extract Safety Score
saf_scores <- data.frame(
  dtmnfr = Safety_Norm$dtmnfr,
  Safety_Index = scale_0_100(pca_safety$ind$coord[, 1])
)

# 4. Extract Mobility Score
mob_scores <- data.frame(
  dtmnfr = Mobility_Norm$dtmnfr,
  Mobility_Index = scale_0_100(pca_mobility$ind$coord[, 1])
)

### Extract the weights from the first principal component
# pca_weights_access <- pca_access$var$coord[, 1] / sum(pca_access$var$coord[, 1])
# pca_weights_afford <- pca_afford$var$coord[, 1] / sum(pca_afford$var$coord[, 1])
# pca_weights_mobility <- pca_mobility$var$coord[, 1] / sum(pca_mobility$var$coord[, 1])
# pca_weights_safety <- pca_safety$var$coord[, 1] / sum(pca_safety$var$coord[, 1])


#### CHECK THE CHAMPIONS INDICATORS

Champions <- function(pca_here, dim = 1, n = 10) {
  # Extract the specific dimension contributions
  contributions <- pca_here$var$contrib[, dim]
  # Sort and take the top n
  # Remember: use TRUE (all caps)
  top_n <- sort(contributions, decreasing = TRUE) |> head(n)
  return(top_n)
}

Champions_access <- Champions(pca_access)
# Champions_afford_pre_nav <- Champions(pca_afford_pre_nav) # or pca_afford_post_nav, depending on which you want to analyze
# Champions_afford_post_nav <- Champions(pca_afford_post_nav)
Champions_mobility <- Champions(pca_mobility)
Champions_safety <- Champions(pca_safety)


# Write the PCA scores to CSV files

write_csv(acc_scores, "/data/IMPT/impt/pca_scores/Accessibility_PCA_Scores.csv")
# write_csv(aff_scores_pre_nav, "Affordability_PCA_Scores_pre_nav.csv")
# write_csv(aff_scores_post_nav, "Affordability_PCA_Scores_post_nav.csv")
write_csv(saf_scores, "/data/IMPT/impt/pca_scores/Safety_PCA_Scores.csv")
write_csv(mob_scores, "/data/IMPT/impt/pca_scores/Mobility_PCA_Scores.csv")

# Write champions scores
write_csv(data.frame(Indicator = names(Champions_access), Contribution = Champions_access), "/data/IMPT/impt/pca_scores/Champions_Accessibility.csv")
# write_csv(data.frame(Indicator = names(Champions_afford_pre_nav), Contribution = Champions_afford_pre_nav), "Champions_Affordability_pre_nav.csv")
# write_csv(data.frame(Indicator = names(Champions_afford_post_nav), Contribution = Champions_afford_post_nav), "Champions_Affordability_post_nav.csv")
write_csv(data.frame(Indicator = names(Champions_mobility), Contribution = Champions_mobility), "/data/IMPT/impt/pca_scores/Champions_Mobility.csv")
write_csv(data.frame(Indicator = names(Champions_safety), Contribution = Champions_safety), "/data/IMPT/impt/pca_scores/Champions_Safety.csv")



#### Join Results scores
Final_Results_Master <- acc_scores |>
  # left_join(aff_scores_pre_nav, by = "dtmnfr") |>
  # left_join(aff_scores_post_nav, by = "dtmnfr") |>
  left_join(afford_singlefare_fin, by = "dtmnfr") |>
  left_join(afford_navegante_fin, by = "dtmnfr") |>
  left_join(mob_scores, by = "dtmnfr") |>
  left_join(saf_scores, by = "dtmnfr")
```


#### PCA computations per mode

The function to rescale is on the previous chunk

```{r}
##################### PCA per mode #####################################

# Perform PCA on the normalized indicators
pca_car_access <- PCA(Accessibility_car_Norm, quali.sup = 1, graph = FALSE)
pca_pt_access <- PCA(Accessibility_pt_Norm, quali.sup = 1, graph = FALSE)
pca_bike_access <- PCA(Accessibility_bike_Norm, quali.sup = 1, graph = FALSE)
pca_walk_access <- PCA(Accessibility_walk_Norm, quali.sup = 1, graph = FALSE)

# pca_car_afford <- PCA(Affordability_car_Norm, quali.sup = 1, graph = FALSE).  NOT ENOUGH INDICATORS
# pca_pt_afford_pre_nav <- PCA(Affordability_pt_Norm_pre_nav, quali.sup = 1, graph = FALSE) NOT ENOUGH INDICATORS
# pca_pt_afford_post_nav <- PCA(Affordability_pt_Norm_post_nav, quali.sup = 1, graph = FALSE) NOT ENOUGH INDICATORS

car_afford <- Affordability_car_Norm |>
  select(dtmnfr, h_transp_inc_car) |>
  rename(Affordability_Index = h_transp_inc_car)
pt_afford_singlefare <- Affordability_pt_Norm_singlefare |>
  select(dtmnfr, h_transp_inc_pt) |>
  rename(Affordability_Index_singlefare = h_transp_inc_pt)
pt_afford_navegante <- Affordability_pt_Norm_navegante |>
  select(dtmnfr, h_transp_inc_pt) |>
  rename(Affordability_Index_navegante = h_transp_inc_pt)

pca_car_mobility <- PCA(Mobility_car_Norm, quali.sup = 1, graph = FALSE)
pca_pt_mobility <- PCA(Mobility_pt_Norm, quali.sup = 1, graph = FALSE)
pca_bike_mobility <- PCA(Mobility_bike_Norm, quali.sup = 1, graph = FALSE)
pca_walk_mobility <- PCA(Mobility_walk_Norm, quali.sup = 1, graph = FALSE)

pca_car_safety <- PCA(Safety_car_Norm, quali.sup = 1, graph = FALSE)
pca_bike_safety <- PCA(Safety_bike_Norm, quali.sup = 1, graph = FALSE)
pca_walk_safety <- PCA(Safety_walk_Norm, quali.sup = 1, graph = FALSE)


####### Obtain the variance explained by PCA1 for each dimension and mode

# Define the list of PCA objects you have created
pca_list <- list(
  Accessibility_GB = pca_access,
  # Affordability_GB_post_nav = pca_afford_post_nav,
  # Affordability_GB_pre_nav = pca_afford_pre_nav,
  Mobility_GB = pca_mobility,
  Safety_GB = pca_safety,
  Accessibility_car = pca_car_access,
  Accessibility_pt = pca_pt_access,
  Accessibility_bike = pca_bike_access,
  Accessibility_walk = pca_walk_access,
  # Affordability_car = car_afford,
  # Affordability_pt_pre_nav = pt_afford_pre_nav,
  # Affordability_pt_post_nav = pt_afford_post_nav,
  Mobility_car = pca_car_mobility,
  Mobility_pt = pca_pt_mobility,
  Mobility_bike = pca_bike_mobility,
  Mobility_walk = pca_walk_mobility,
  Safety_car = pca_car_safety,
  Safety_bike = pca_bike_safety,
  Safety_walk = pca_walk_safety
)

# Function to extract variance and variable correlations
get_pca_diagnostics <- function(pca_obj, name) {
  # 1. Get Variance of PC1
  var_pc1 <- pca_obj$eig[1, 2]
  # 2. Get Correlations of variables with PC1
  cors <- as.data.frame(pca_obj$var$cor[, 1])
  colnames(cors) <- "Correlation_PC1"
  cors$Variable <- rownames(cors)
  cors$Dimension_Mode <- name
  cors$PC1_Variance_Percent <- var_pc1
  return(cors)
}

# Apply the function to all PCAs and combine into one table
pca_variance_report <- do.call(rbind, lapply(names(pca_list), function(x) {
  get_pca_diagnostics(pca_list[[x]], x)
}))

# Reorder columns for readability
pca_variance_report <- pca_variance_report[, c("Dimension_Mode", "PC1_Variance_Percent", "Variable", "Correlation_PC1")]


######## Write the results to export PCA variances

write_csv(pca_variance_report, "/data/IMPT/impt/pca_scores/PCA_Variance_and_Correlations.csv")

pca_summary_table <- data.frame(
  Dimension_Mode = names(pca_list),
  PC1_Variance_Percent = sapply(pca_list, function(x) x$eig[1, 2]),
  Eigenvalue_PC1 = sapply(pca_list, function(x) x$eig[1, 1])
)

# 3. View the clean summary
print(pca_summary_table)

#### Export CSV summary

write_csv(pca_summary_table, "/data/IMPT/impt/pca_scores/PCA_Variance_Table.csv")


# Check the direction of the first variable for each PCA
# 1. Run the loop and store it in an object
direction_results <- do.call(rbind, lapply(names(pca_list), function(x) {
  # Get the correlations for the 1st component
  cors <- pca_list[[x]]$var$cor[, 1]
  # Find the variable with the strongest relationship to PC1
  top_idx <- which.max(abs(cors))
  top_var <- names(cors)[top_idx]
  cor_val <- cors[top_idx]
  data.frame(
    Dimension = x,
    Top_Variable = top_var,
    Correlation = round(cor_val, 3)
  )
}))

# 2. Explicitly print it to the console
print(direction_results)

# 1. Extract Accessibility Score
acc_car_scores <- data.frame(
  dtmnfr = Accessibility_car_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(as.numeric(pca_car_access$ind$coord[, 1]))
)

acc_pt_scores <- data.frame(
  dtmnfr = Accessibility_pt_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(as.numeric(pca_pt_access$ind$coord[, 1]))
)

acc_bike_scores <- data.frame(
  dtmnfr = Accessibility_bike_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(as.numeric(pca_bike_access$ind$coord[, 1]))
)

acc_walk_scores <- data.frame(
  dtmnfr = Accessibility_walk_Norm$dtmnfr,
  Accessibility_Index = scale_0_100(as.numeric(pca_walk_access$ind$coord[, 1]))
)

# 2. Extract Affordability Score
aff_car_scores <- car_afford
aff_pt_scores_singlefare <- pt_afford_singlefare
aff_pt_scores_navegante <- pt_afford_navegante


# # 3. Extract Safety Score
# saf_car_scores <- data.frame(dtmnfr = Safety_car_Norm$dtmnfr,
#                              Safety_Index = scale_0_100(pca_car_safety$ind$coord[, 1]))
# saf_bike_scores <- data.frame(dtmnfr = Safety_bike_Norm$dtmnfr,
#                               Safety_Index = scale_0_100(pca_bike_safety$ind$coord[, 1]))
# saf_walk_scores <- data.frame(dtmnfr = Safety_walk_Norm$dtmnfr,
#                               Safety_Index = scale_0_100(pca_walk_safety$ind$coord[, 1]))

# 4. Extract Mobility Score
mob_car_scores <- data.frame(
  dtmnfr = Mobility_car_Norm$dtmnfr,
  Mobility_Index = scale_0_100(pca_car_mobility$ind$coord[, 1])
)

######### Safety per mode using avg###########

saf_car_scores <- Safety_car_Norm |>
  mutate(indice_gravidade_carro = 100 - indice_gravidade_carro) |>
  mutate(Safety_Index = indice_gravidade_carro * 0.5 + veh_motorizado * 0.5) |>
  select(dtmnfr, Safety_Index)
saf_bike_scores <- Safety_bike_Norm |>
  mutate(indice_gravidade_bicicleta = 100 - indice_gravidade_bicicleta) |>
  mutate(Safety_Index = indice_gravidade_bicicleta * 0.5 + veh_bicicleta * 0.5) |>
  select(dtmnfr, Safety_Index)
saf_walk_scores <- Safety_walk_Norm |>
  mutate(indice_gravidade_peoes = 100 - indice_gravidade_peoes) |>
  mutate(Safety_Index = indice_gravidade_peoes * 0.5 + veh_peoes * 0.5) |>
  select(dtmnfr, Safety_Index)

mob_pt_scores <- data.frame(
  dtmnfr = Mobility_pt_Norm$dtmnfr,
  Mobility_Index = scale_0_100(pca_pt_mobility$ind$coord[, 1])
)
mob_bike_scores <- data.frame(
  dtmnfr = Mobility_bike_Norm$dtmnfr,
  Mobility_Index = scale_0_100(pca_bike_mobility$ind$coord[, 1])
)
mob_walk_scores <- data.frame(
  dtmnfr = Mobility_walk_Norm$dtmnfr,
  Mobility_Index = scale_0_100(pca_walk_mobility$ind$coord[, 1])
)

#### Join Results scores

Final_Results_Master_car <- acc_car_scores |>
  left_join(car_afford, by = "dtmnfr") |>
  left_join(mob_car_scores, by = "dtmnfr") |>
  left_join(saf_car_scores, by = "dtmnfr")

Final_Results_Master_pt <- acc_pt_scores |>
  left_join(pt_afford_navegante, by = "dtmnfr") |>
  left_join(pt_afford_singlefare, by = "dtmnfr") |>
  left_join(mob_pt_scores, by = "dtmnfr")

Final_Results_Master_bike <- acc_bike_scores |>
  left_join(mob_bike_scores, by = "dtmnfr") |>
  left_join(saf_bike_scores, by = "dtmnfr")

Final_Results_Master_walk <- acc_walk_scores |>
  left_join(mob_walk_scores, by = "dtmnfr") |>
  left_join(saf_walk_scores, by = "dtmnfr")


# Extract the weights from the first principal component
# pca_weights_access <- pca_access$var$coord[, 1] / sum(pca_access$var$coord[, 1])
# pca_weights_afford <- pca_afford$var$coord[, 1] / sum(pca_afford$var$coord[, 1])
# pca_weights_mobility <- pca_mobility$var$coord[, 1] / sum(pca_mobility$var$coord[, 1])
# pca_weights_safety <- pca_safety$var$coord[, 1] / sum(pca_safety$var$coord[, 1])


# Write the PCA scores to CSV files

write_csv(acc_car_scores, "/data/IMPT/impt/pca_scores/Accessibility_PCA_car_Scores.csv")
write_csv(acc_pt_scores, "/data/IMPT/impt/pca_scores/Accessibility_PCA_pt_Scores.csv")
write_csv(acc_bike_scores, "/data/IMPT/impt/pca_scores/Accessibility_PCA_bike_Scores.csv")
write_csv(acc_walk_scores, "/data/IMPT/impt/pca_scores/Accessibility_PCA_walk_Scores.csv")

write_csv(aff_car_scores, "/data/IMPT/impt/pca_scores/Affordability_car_Scores.csv")
write_csv(aff_pt_scores_singlefare, "/data/IMPT/impt/pca_scores/Affordability_pt_Scores_singlefare.csv")
write_csv(aff_pt_scores_navegante, "/data/IMPT/impt/pca_scores/Affordability_pt_Scores_navegante.csv")

write_csv(saf_car_scores, "/data/IMPT/impt/pca_scores/Safety_car_Scores.csv")
write_csv(saf_bike_scores, "/data/IMPT/impt/pca_scores/Safety_bike_Scores.csv")
write_csv(saf_walk_scores, "/data/IMPT/impt/pca_scores/Safety_walk_Scores.csv")

write_csv(mob_car_scores, "/data/IMPT/impt/pca_scores/Mobility_PCA_car_Scores.csv")
write_csv(mob_pt_scores, "/data/IMPT/impt/pca_scores/Mobility_PCA_pt_Scores.csv")
write_csv(mob_bike_scores, "/data/IMPT/impt/pca_scores/Mobility_PCA_bike_Scores.csv")
write_csv(mob_walk_scores, "/data/IMPT/impt/pca_scores/Mobility_PCA_walk_Scores.csv")
```


### Step 4: Compute the overall IMPT score

```{r}
# ### EQUAL WEIGHTS
# Results_Equal_Weights <- Results_Equal_Weights |>
#   mutate(
#     # Do the shifting and the math all in one go
#     IMPT_Equal_Weights_geom = ((Accessibility + 0.01) *
#                                  (Affordability + 0.01) *
#                                  (Mobility + 0.01) *
#                                  (Safety  + 0.01)
#     )^(1 / 4) # weight equal
#   ) |>
#   mutate(IMPT_Equal_Weights_avg = (Accessibility + Affordability + Mobility + Safety) / 4)


####### Compute overall IMPT score using PCA weights + AVG and geometric mean
Final_Results_Master <- Final_Results_Master |>
  mutate( # Adding +1 ensures no zeros, then we subtract it at the end
    IMPT_score_pca_geom_navegante = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Affordability_Index_navegante + 1) * (Safety_Index + 1))^(1 / 4) - 1
  ) |>
  mutate(IMPT_score_pca_geom_singlefare = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Affordability_Index_singlefare + 1) * (Safety_Index + 1))^(1 / 4) - 1) |>
  mutate(IMPT_score_pca_avg_singlefare = (Accessibility_Index + Mobility_Index + Affordability_Index_singlefare + Safety_Index) / 4) |>
  mutate(IMPT_score_pca_avg_navegante = (Accessibility_Index + Mobility_Index + Affordability_Index_navegante + Safety_Index) / 4)


# Compute overall IMPT with entropy weights
# Assuming 'entropy_weights' is a vector of weights derived from the entropy method for each dimension

compute_entropy_weights <- function(df) {
  # Add 0.0001 to avoid log(0)
  df_shifted <- df + 0.0001
  # Calculate proportions
  p_ij <- sweep(df_shifted, 2, colSums(df_shifted), "/")
  # Entropy calculation
  n <- nrow(df)
  k <- 1 / log(n)
  # Calculate e_j, using 0 for cases where p_ij is 0 to avoid log issues
  e_j <- -k * colSums(p_ij * log(p_ij + 1e-10))
  d_j <- 1 - e_j
  w_j <- d_j / sum(d_j)
  return(w_j)
}

# 1. dataframe containing only the scores you want to weight
scores_for_entropy1 <- Final_Results_Master |>
  select(Accessibility_Index, Affordability_Index_navegante, Mobility_Index, Safety_Index)
scores_for_entropy2 <- Final_Results_Master |>
  select(Accessibility_Index, Affordability_Index_singlefare, Mobility_Index, Safety_Index)

Entropy_weights_pca_navegante <- compute_entropy_weights(scores_for_entropy1)
Entropy_weights_pca_nonav <- compute_entropy_weights(scores_for_entropy2)

# 3. Compute the Index and final results

# Using a +1 shift is very common for 0-100 scales
Final_Results_Master <- Final_Results_Master |>
  mutate(
    IMPT_entropy_pca_navegante = (
      (Accessibility_Index + 1)^Entropy_weights_pca_navegante[1] *
        (Affordability_Index_navegante + 1)^Entropy_weights_pca_navegante[2] *
        (Mobility_Index + 1)^Entropy_weights_pca_navegante[3] *
        (Safety_Index + 1)^Entropy_weights_pca_navegante[4]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  ) |>
  mutate(
    IMPT_entropy_pca_singlefare = (
      (Accessibility_Index + 1)^Entropy_weights_pca_nonav[1] *
        (Affordability_Index_singlefare + 1)^Entropy_weights_pca_nonav[2] *
        (Mobility_Index + 1)^Entropy_weights_pca_nonav[3] *
        (Safety_Index + 1)^Entropy_weights_pca_nonav[4]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  )


######### Save the final results
# write_csv(Results_Equal_Weights, "/data/IMPT/impt/results/IMPT_Equal_Weights.csv")
write_csv(Final_Results_Master, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_freguesia.csv")
```


#### Compute IMPT per mode

```{r}
################################################### Compute overall IMPT score using PCA weights + AVG and geometric mean
Final_Results_Master_car <- Final_Results_Master_car |>
  mutate( # Adding +1 ensures no zeros, then we subtract it at the end
    IMPT_score_pca_geom_car = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Affordability_Index + 1) * (Safety_Index + 1))^(1 / 4) - 1
  ) |>
  mutate(IMPT_score_pca_avg_car = ((Accessibility_Index) + (Mobility_Index) + (Affordability_Index) + (Safety_Index)) / 4)


Final_Results_Master_pt <- Final_Results_Master_pt |>
  mutate( # Adding +1 ensures no zeros, then we subtract it at the end
    IMPT_score_pca_geom_pt_singlefare = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Affordability_Index_singlefare + 1))^(1 / 3) - 1
  ) |>
  mutate(
    IMPT_score_pca_geom_pt_navegante = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Affordability_Index_navegante + 1))^(1 / 3) - 1
  ) |>
  mutate(
    IMPT_score_pca_avg_pt_singlefare = ((Accessibility_Index) + (Mobility_Index) + (Affordability_Index_singlefare)) / 3
  ) |>
  mutate(
    IMPT_score_pca_avg_pt_navegante = ((Accessibility_Index) + (Mobility_Index) + (Affordability_Index_navegante)) / 3
  )

Final_Results_Master_bike <- Final_Results_Master_bike |>
  mutate( # Adding +1 ensures no zeros, then we subtract it at the end
    IMPT_score_pca_geom_bike = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Safety_Index + 1))^(1 / 3) - 1
  ) |>
  mutate(
    IMPT_score_pca_avg_bike = ((Accessibility_Index) + (Mobility_Index) + (Safety_Index)) / 4
  )

Final_Results_Master_walk <- Final_Results_Master_walk |>
  mutate( # Adding +1 ensures no zeros, then we subtract it at the end
    IMPT_score_pca_geom_walk = ((Accessibility_Index + 1) * (Mobility_Index + 1) * (Safety_Index + 1))^(1 / 3) - 1
  ) |>
  mutate(
    IMPT_score_pca_avg_walk = ((Accessibility_Index) + (Mobility_Index) + (Safety_Index)) / 3
  )

#### Compute overall IMPT with entropy weights

# 1. dataframe containing only the scores you want to weight
scores_for_entropy_car <- Final_Results_Master_car |>
  select(Accessibility_Index, Affordability_Index, Mobility_Index, Safety_Index)
Entropy_weights_pca_car <- compute_entropy_weights(scores_for_entropy_car)

scores_for_entropy_pt_navegante <- Final_Results_Master_pt |>
  select(Accessibility_Index, Affordability_Index_navegante, Mobility_Index)
scores_for_entropy_pt_singlefare <- Final_Results_Master_pt |>
  select(Accessibility_Index, Affordability_Index_singlefare, Mobility_Index)
Entropy_weights_pca_pt_navegante <- compute_entropy_weights(scores_for_entropy_pt_navegante)
Entropy_weights_pca_pt_singlefare <- compute_entropy_weights(scores_for_entropy_pt_singlefare)

scores_for_entropy_bike <- Final_Results_Master_bike |>
  select(Accessibility_Index, Mobility_Index, Safety_Index)
Entropy_weights_pca_bike <- compute_entropy_weights(scores_for_entropy_bike)

scores_for_entropy_walk <- Final_Results_Master_walk |>
  select(Accessibility_Index, Mobility_Index, Safety_Index)
Entropy_weights_pca_walk <- compute_entropy_weights(scores_for_entropy_walk)

# 3. Compute the Index and final results

# Using a +1 shift is very common for 0-100 scales
Final_Results_Master_car <- Final_Results_Master_car |>
  mutate(
    IMPT_entropy_pca_car = (
      (Accessibility_Index + 1)^Entropy_weights_pca_car[1] *
        (Affordability_Index + 1)^Entropy_weights_pca_car[2] *
        (Mobility_Index + 1)^Entropy_weights_pca_car[3] *
        (Safety_Index + 1)^Entropy_weights_pca_car[4]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  )

Final_Results_Master_pt <- Final_Results_Master_pt |>
  mutate(
    IMPT_entropy_pca_pt_navevante = (
      (Accessibility_Index + 1)^Entropy_weights_pca_pt_navegante[1] *
        (Affordability_Index_navegante + 1)^Entropy_weights_pca_pt_navegante[2] *
        (Mobility_Index + 1)^Entropy_weights_pca_pt_navegante[3]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  ) |>
  mutate(
    IMPT_entropy_pca_pt_singlefare = (
      (Accessibility_Index + 1)^Entropy_weights_pca_pt_singlefare[1] *
        (Affordability_Index_singlefare + 1)^Entropy_weights_pca_pt_singlefare[2] *
        (Mobility_Index + 1)^Entropy_weights_pca_pt_singlefare[3]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  )

Final_Results_Master_bike <- Final_Results_Master_bike |>
  mutate(
    IMPT_entropy_pca_bike = (
      (Accessibility_Index + 1)^Entropy_weights_pca_bike[1] *
        (Mobility_Index + 1)^Entropy_weights_pca_bike[2] *
        (Safety_Index + 1)^Entropy_weights_pca_bike[3]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  )

Final_Results_Master_walk <- Final_Results_Master_walk |>
  mutate(
    IMPT_entropy_pca_walk = (
      (Accessibility_Index + 1)^Entropy_weights_pca_walk[1] *
        (Mobility_Index + 1)^Entropy_weights_pca_walk[2] *
        (Safety_Index + 1)^Entropy_weights_pca_walk[3]
    ) - 1 # Subtract the 1 at the end to keep the scale similar
  )


######### Save the final results
write_csv(Final_Results_Master_car, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_car_freguesia.csv")
write_csv(Final_Results_Master_pt, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_pt_freguesia.csv")
write_csv(Final_Results_Master_bike, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_bike_freguesia.csv")
write_csv(Final_Results_Master_walk, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_walk_freguesia.csv")
```

## Export for other levels

```{r}
## Municipality
# export at municipality level, weighted mean with population
census24_fregmun_pop <- read_csv("useful_data/census24_fregmun.csv") |>
  select(freg_id, mun_id, population) |>
  mutate(dtmnfr = as.integer(freg_id))

Final_Results_Master_municipio <- Final_Results_Master |>
  left_join(census24_fregmun_pop, by = "dtmnfr") |>
  group_by(mun_id) |>
  # weighted mean with population as weights
  summarize_at(vars(-dtmnfr, -freg_id, -population), ~ weighted.mean(., w = population, na.rm = TRUE))
Final_Results_Master_car_municipio <- Final_Results_Master_car |>
  left_join(census24_fregmun_pop, by = "dtmnfr") |>
  group_by(mun_id) |>
  # weighted mean with population as weights
  summarize_at(vars(-dtmnfr, -freg_id, -population), ~ weighted.mean(., w = population, na.rm = TRUE))
Final_Results_Master_pt_municipio <- Final_Results_Master_pt |>
  left_join(census24_fregmun_pop, by = "dtmnfr") |>
  group_by(mun_id) |>
  # weighted mean with population as weights
  summarize_at(vars(-dtmnfr, -freg_id, -population), ~ weighted.mean(., w = population, na.rm = TRUE))
Final_Results_Master_walk_municipio <- Final_Results_Master_walk |>
  left_join(census24_fregmun_pop, by = "dtmnfr") |>
  group_by(mun_id) |>
  # weighted mean with population as weights
  summarize_at(vars(-dtmnfr, -freg_id, -population), ~ weighted.mean(., w = population, na.rm = TRUE))
Final_Results_Master_bike_municipio <- Final_Results_Master_bike |>
  left_join(census24_fregmun_pop, by = "dtmnfr") |>
  group_by(mun_id) |>
  # weighted mean with population as weights
  summarize_at(vars(-dtmnfr, -freg_id, -population), ~ weighted.mean(., w = population, na.rm = TRUE))

write_csv(Final_Results_Master_municipio, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_municipio.csv")
write_csv(Final_Results_Master_car_municipio, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_car_municipio.csv")
write_csv(Final_Results_Master_pt_municipio, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_pt_municipio.csv")
write_csv(Final_Results_Master_walk_municipio, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_walk_municipio.csv")
write_csv(Final_Results_Master_bike_municipio, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_bike_municipio.csv")


## Grid level
# via grid_freg_mun lookup (generalised assumption).
grid_freg_mun <- read.csv("useful_data/grid_nuts.csv") |>
  select(grid_id, freg_id) |>
  mutate(freg_id = as.integer(freg_id))

Final_Results_Master_grid <- grid_freg_mun |>
  left_join(Final_Results_Master, by = c("freg_id" = "dtmnfr")) |>
  select(-freg_id)
Final_Results_Master_grid_car <- grid_freg_mun |>
  left_join(Final_Results_Master_car, by = c("freg_id" = "dtmnfr")) |>
  select(-freg_id)
Final_Results_Master_grid_pt <- grid_freg_mun |>
  left_join(Final_Results_Master_pt, by = c("freg_id" = "dtmnfr")) |>
  select(-freg_id)
Final_Results_Master_grid_walk <- grid_freg_mun |>
  left_join(Final_Results_Master_walk, by = c("freg_id" = "dtmnfr")) |>
  select(-freg_id)
Final_Results_Master_grid_bike <- grid_freg_mun |>
  left_join(Final_Results_Master_bike, by = c("freg_id" = "dtmnfr")) |>
  select(-freg_id)


write.csv(Final_Results_Master_grid, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_grid.csv", row.names = FALSE)
write.csv(Final_Results_Master_grid_car, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_car_grid.csv", row.names = FALSE)
write.csv(Final_Results_Master_grid_pt, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_pt_grid.csv", row.names = FALSE)
write.csv(Final_Results_Master_grid_bike, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_bike_grid.csv", row.names = FALSE)
write.csv(Final_Results_Master_grid_walk, "/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_walk_grid.csv", row.names = FALSE)
```


# Visualize

```{r}
library(mapview)
freguesias # Read from 00_data_load.R

# DATA_LOCATION = "https://impt.server.ushift.pt" # When running locally, get data from remote server
# API_KEY = Sys.getenv("IMPT_DATA_KEY") # Set it using usethis::edit_r_environ(), followed by CTRL+F10
#
# IMPT_URL = function(path) {
#   # If does not start with "/", add it
#   if (!startsWith(path, "/")) {
#     path = paste0("/", path)
#   }
#   local_path = sprintf("%s%s", DATA_LOCATION, path)
#   # If data location starts with "http", add api key to url
#   if (startsWith(DATA_LOCATION, "http")) {
#     # If API_KEY empty or not defined, throw error
#     if (API_KEY == "") {
#       stop("IMPT_DATA_KEY env var is not defined. Please set it using usethis::edit_r_environ() and restart R.")
#     }
#     return(sprintf("%s%s?key=%s", DATA_LOCATION, path, API_KEY))
#   } else {
#     # If local folder does not exist, create, recursively if needed
#     if (!dir.exists(dirname(local_path))) {
#       dir.create(dirname(local_path), recursive = TRUE)
#     }
#   }
#   # Otherwise, return local path
#   return(local_path)
# }


# IMPT PCA and Entropy scores
# impt_pca = impt_read("results_aggregated/20260330/IMPT_PCA_and_Entropy_Scores.csv")
impt_pca <- read.csv("/data/IMPT/impt/results/IMPT_PCA_and_Entropy_Scores_freguesia.csv")
impt_pca_sf <- freguesias |>
  select(-nuts2, -area_ha) |>
  mutate(dtmnfr = as.integer(dtmnfr)) |>
  left_join(
    impt_pca |>
      rename(
        Accessibility_PCA = Accessibility_Index,
        Affordability_PCA_prenav = Affordability_Index_singlefare,
        Affordability_PCA_postnav = Affordability_Index_navegante,
        Mobility_PCA = Mobility_Index,
        Safety_PCA = Safety_Index
      )
  ) |>
  # Round PCA scores to 2 decimals
  mutate(
    across(starts_with("Accessibility"), ~ round(., 2)),
    across(starts_with("Affordability"), ~ round(., 2)),
    across(starts_with("Mobility"), ~ round(., 2)),
    across(starts_with("Safety"), ~ round(., 2)),
    across(starts_with("IMPT"), ~ round(., 2))
  ) |>
  # Columns order
  select(
    dtmnfr, freguesia, municipio,
    starts_with("Accessibility"), starts_with("Affordability"), starts_with("Mobility"), starts_with("Safety"),
    starts_with("IMPT")
  )
# View(impt_pca_sf |> st_drop_geometry())

mapview(impt_pca_sf, zcol = "Accessibility_PCA", layer.name = "PCA Accessibility", hide = TRUE) +
  mapview(impt_pca_sf, zcol = "Affordability_PCA_prenav", layer.name = "PCA Affordability - single fare", hide = TRUE) +
  mapview(impt_pca_sf, zcol = "Affordability_PCA_postnav", layer.name = "PCA Affordability - navegante", hide = TRUE) +
  mapview(impt_pca_sf, zcol = "Mobility_PCA", layer.name = "PCA Mobility", hide = TRUE) +
  mapview(impt_pca_sf, zcol = "Safety_PCA", layer.name = "PCA Safety", hide = TRUE) +

  mapview(impt_pca_sf, zcol = "IMPT_score_pca_geom_navegante", layer.name = "IMPT PCA Geom - navegante", hide = TRUE) +
  mapview(impt_pca_sf, zcol = "IMPT_score_pca_geom_singlefare", layer.name = "IMPT PCA Geom - single fare", hide = TRUE) +
  mapview(impt_pca_sf, zcol = "IMPT_score_pca_avg_navegante", layer.name = "IMPT PCA Avg - navegante", hide = TRUE) +
  mapview(impt_pca_sf, zcol = "IMPT_score_pca_avg_singlefare", layer.name = "IMPT PCA Avg - single fare", hide = TRUE) +
  mapview(impt_pca_sf, zcol = "IMPT_entropy_pca_navegante", layer.name = "IMPT PCA Entropy - navegante", hide = TRUE) +
  mapview(impt_pca_sf, zcol = "IMPT_entropy_pca_singlefare", layer.name = "IMPT PCA Entropy - single fare", hide = TRUE)
```

